---
title: "The software stack"
subtitle: "What R, Stan, `brms` and `bayesnec` each do, and why all four are needed"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

Fitting a concentration-response model in this course sets four pieces of
software working together, and a failure in any of them surfaces as an error in
whichever one you were typing into. Knowing what each layer does makes those
errors readable, and it explains several things about the workflow that are
otherwise arbitrary: why the first model of a session is slow, why a C++
compiler has to be installed to fit a curve, and why a fitted object is large.

## The working environment

R is the language. Positron is the editor used throughout the course: an
environment for writing and running R, with the script you are editing in one
pane and a live R session, the console, in another. Code is sent from the script
to the console a line or a block at a time, so the script is the record of what
was done and the console is where it happens.

Two habits matter more here than in ordinary R work.

**Work in a project.** A project fixes the working directory, so relative paths
like `read.csv("mydata.csv")` resolve the same way for everyone. Without one, the
same script behaves differently on two machines for reasons that are tedious to
diagnose.

**Do not save the workspace.** R offers to store the objects in memory when a
session ends and to restore them when it starts. Accept that and you will
eventually be working with a fitted model whose provenance you cannot reconstruct
— fitted under code you have since edited, indistinguishable from a current one.
Start every session empty and re-run the script. The exception is a Bayesian fit,
which is expensive enough to save deliberately to a file, covered in module 2.

## The four layers

```{mermaid}
%%| label: fig-stack
%%| fig-cap: "Each layer is called by the one above it."
flowchart TD
    A["<b>bayesnec</b><br/>concentration-response models,<br/>priors, toxicity estimates"]
    B["<b>brms</b><br/>R formulas to Stan code"]
    C["<b>cmdstanr</b> or <b>rstan</b><br/>the interface to Stan"]
    D["<b>Stan</b><br/>the sampler, written in C++"]
    E["<b>C++ compiler</b><br/>Rtools, Xcode CLT, or gcc"]
    A --> B --> C --> D --> E
```

**Stan** is the engine. It is a separate piece of software, not an R package,
that performs Bayesian inference by Hamiltonian Monte Carlo. A Stan program
describes a statistical model; Stan translates it to C++, compiles it into an
executable, and runs that executable to draw samples from the posterior. This is
the source of the compile step: a Stan model is a program that has to be built
before it can run, which is why a C++ compiler is a requirement and why the first
fit of a session takes noticeably longer than later ones.

**`cmdstanr` and `rstan`** are two R interfaces to Stan, and `brms` can use
either. This course uses `cmdstanr`, which drives a normal CmdStan installation
as a separate process. `rstan` embeds Stan inside R instead, which couples the
Stan version to the package version and has historically been the more fragile
arrangement on Windows. `rstan` still has to be installed, because `brms` imports
it, but it is not the sampler here. The backend is selected once:

```{r}
#| label: backend
#| eval: false
options(brms.backend = "cmdstanr")
```

**`brms`** [@Burkner2017; @Burkner2018] removes the need to write Stan code. It
takes a model written in R's formula syntax, close to that of `lme4`, and
generates the Stan program for it. It supports a wide range of response
distributions and model structures — non-linear and smooth terms, multilevel
structure, censoring, distributional regression — and provides the methods for
checking a fit once it exists.

**`bayesnec`** [@Fisher2020] is what this course is about. `brms` will fit a
non-linear model, but only if you supply the equation, sensible starting values
and priors for every parameter. For concentration-response work that is the
barrier: the models are non-linear, the priors are not obvious, and a poorly
initialised non-linear fit fails in ways that need specialist reading. This is a
substantial part of why Bayesian methods are not widely used for *NEC* estimation
in ecotoxicology, where specialist statistical support is rarely available.

`bayesnec` takes a short formula naming a concentration-response equation, and
from that plus the data it constructs the full non-linear `brms` formula,
generates priors and initial values, chooses a statistical distribution
appropriate to the response, fits the model, and extracts toxicity estimates from
it. Module 2 does exactly this in one line.

## Where bayesnec came from

The package began as `jagsNEC` [@Fisher2020], an implementation of the *NEC*
model of @Fox2010 using JAGS. It was rewritten on `brms` and Stan, and
generalised well beyond the original: the supported response distributions now
include the beta-binomial alongside the Gaussian, Poisson, binomial, gamma,
negative binomial and beta of the original, and any `brms` family can in
principle be added.

The model set was also extended. Alongside the threshold models that estimate an
*NEC* directly, it now includes smooth concentration-response equations of the
kind used in frequentist packages such as `drc` [@Ritz2015] — four-parameter
logistic and Weibull curves among them — which model the response as a smooth
function of concentration rather than with a breakpoint. Module 3 sets out the
full set and what each one supports.

## Versions

The versions in use when a model is fitted are part of the result, because
defaults and estimators change between releases. Recording them costs one line:

```{r}
#| label: versions
#| comment: ""
for (p in c("bayesnec", "brms", "cmdstanr", "rstan", "posterior")) {
  cat(sprintf("%-10s %s\n", p, as.character(packageVersion(p))))
}
cat(sprintf("%-10s %s\n", "R", getRversion()))
```

This course is built against the development version of `bayesnec`, which is
ahead of the release on CRAN. The setup instructions name the version to install.

## Getting help, and reporting problems

The `bayesnec` reference site at
[open-aims.github.io/bayesnec](https://open-aims.github.io/bayesnec/) documents
every exported function and carries articles on single-model and multi-model
usage, the model set, priors, and comparing posterior predictions. This course
covers much of the same ground in a different order.

The package is under active development and feedback is wanted. Problems go to
the [issue tracker](https://github.com/open-AIMS/bayesnec/issues).

A report is only actionable with a self-contained reproducible example: code and
data that someone else can run to see the same behaviour. This matters
particularly for a fit that fails to converge, because the cause is usually
specific to the data. Without a reproducible example a convergence problem cannot
be diagnosed by anyone who does not have your dataset.
