---
title: "The software stack"
subtitle: "What R, Stan, `brms` and `bayesnec` each do, and why all four are needed"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

Fitting a concentration-response model in this course requires several pieces of
software to working together, and a failure in any of them will mean the model cannot be fitted. 
Knowing what each layer does makes those
errors easier to understand, and it explains several things about the workflow that are
otherwise arbitrary: why the first model of a session is slow, why a C++
compiler has to be installed to fit a curve, and why a fitted object is large.

## The working environment

bayesnec is written in the R statistical programming language. While it is possible
to use R from some things as a stand alone program, there are additional open-source 
third-party software that make R more intuitive and easy to use. Almost no one uses R by itself.
The course uses the Positron integrated development environment (IDE) for R, 
which is a free and open-source software that provides a user-friendly interface to R.
It is available for Windows, MacOS, and Linux operating systems, and is the more 
recently preferred IDE for R
and will likely replace the popular RStudio IDE for most users in the future.

A new session starts empty.

![Positron with a new R session. The console reports which R started, and the
interpreter in use is named at the top right with the open project beside
it.](images/positron_R.png){#fig-positron-new}


![The same window after running a short script. The console echoes the lines it
was sent, the Variables pane lists the objects that now exist and what they
contain, and the plot appears in the Plots pane.](images/positron_R2.png){#fig-positron-run}

Four parts of that window are worth knowing before the first model is fitted.

**The editor**, top left, holds the script. Run the current line or selection
with `Ctrl`+`Enter` (`Cmd`+`Enter` on macOS) and it is sent to the console below.

**The console**, bottom left, is the running R session. It echoes what it was
sent, so @fig-positron-run shows the same four lines in both panes.

**Variables**, top right, lists every object in the session with its type and
contents — `x` as an integer vector of ten values and `y` as a double in the
figure. It is the quickest way to see whether an object is what you expected,
which matters later when a fitted model is a deeply nested list rather than a
vector.

**Plots**, bottom right, collects the figures produced, with arrows to step back
through earlier ones and a control to save the current figure.

Probably the most important thing is the editor pane, because it is where the script is written. 
The script is the record of what was done, and it can be re-run from the top to reproduce the results. 

The console is where the code is actually run, and it is where the output of the code is displayed. 

The editor and console are linked: 
Code is sent from the script to the console a line or a block at a time, rather
than typed into the console directly. 

This scripting workflow is a better way of working in many ways than most pre-packages software, 
where the user clicks buttons and the software does things behind the scenes. 

Having a script means that you have a permanent record of everything that was done 
in the analyis, so that the results can be completely reproduced, which is something that
is very important in environmental sciences (or any science).

The console is where the code is actually run, and it is where the output of the code is displayed.

The interpreter (in our case R version) in use is named at the top right of the window, 
with the open project beside it. Check it when something is missing that you know you
installed: more than one R on a machine is common, and each has its own library,
so a package installed under one is absent under another. You can also run other software, 
such as python, from inside positrons console, but that is not needed in this course.

Two habits matter more here than in ordinary R work, althought I would apply them
to any R projects I am working on.

**Work in a project.** A project fixes the working directory, so relative paths
like `read.csv("mydata.csv")` resolve the same way for everyone. Without one, the
same script behaves differently on two machines for reasons that are tedious to
diagnose.

**Do not save the workspace.** R offers to store the objects in memory when a
session ends and to restore them when it starts. Accept that and you will
eventually be working with a fitted model whose provenance you cannot reconstruct
— fitted under code you have since edited, indistinguishable from a current one.
Start every session empty and re-run the script. The exception is a Bayesian fit,
which is expensive enough to save deliberately to a file, covered in module 2.

## The dependency chain

```{mermaid}
%%| label: fig-stack
%%| fig-cap: "Each layer is called by the one above it."
flowchart TD
    A["<b>bayesnec</b><br/>concentration-response models,<br/>priors, toxicity estimates"]
    B["<b>brms</b><br/>R formulas to Stan code"]
    C["<b>cmdstanr</b> or <b>rstan</b><br/>the interface to Stan"]
    D["<b>Stan</b><br/>the sampler, written in C++"]
    E["<b>C++ compiler</b><br/>Rtools, Xcode CLT, or gcc"]
    A --> B --> C --> D --> E
```

**Stan** is the engine. It is a separate piece of software, not an R package,
that performs Bayesian inference by Hamiltonian Monte Carlo. A Stan program
describes a statistical model; Stan translates it to C++, compiles it into an
executable, and runs that executable to draw samples from the posterior. This is
the source of the compile step: a Stan model is a program that has to be built
before it can run, which is why a C++ compiler is a requirement and why the first
fit of a session takes noticeably longer than later ones.

This may seem very opaque of you are not familiar with Bayesian inference, 
but we will cover more detail about Bayesian inference and how Stan works in module 
6.

**`cmdstanr` and `rstan`** are two R interfaces to Stan, and `brms` can use
either. This course uses `cmdstanr`, which drives a normal CmdStan installation
as a separate process. `rstan` embeds Stan inside R instead, which couples the
Stan version to the package version and has historically been the more fragile
arrangement on Windows. `rstan` still has to be installed, because `brms` imports
it, but it is not the sampler here. The backend is selected once:

```{r}
#| label: backend
#| eval: false
options(brms.backend = "cmdstanr")
```

**`brms`** [@Burkner2017; @Burkner2018] removes the need to write Stan code,
which is quite complicated and beyond the capability of most ecotoxicologists to implement in practice. 

Relatively complex models can be fit using `brms`, using a relatively straightforward formula syntax. 
A typical model specification might be
`response ~ predictor1 + predictor2 + (1 | group)`,
which is similar to the syntax used in many widely used R packages, such as `lme4`, 
for example.

`brms` supports a wide range of response
distributions and model structures — non-linear and smooth terms, multilevel
structure, censoring, distributional regression — and provides the methods for
checking a fit once it exists. Why these are useful will be discussed in later modules.

**`bayesnec`** [@Fisher2020] is the software package this course is about. `brms` will fit a
non-linear model, but only if you supply the equation, sensible starting values
and priors for every parameter. For concentration-response work that is the
barrier: the models are non-linear, the priors are not obvious, and a poorly
initialised non-linear fit fails in ways that need relatively advanced statistical 
understanding in order to resolve. This is a
substantial part of why Bayesian methods are not widely used for *NEC* estimation
in ecotoxicology, where specialist statistical support is rarely or infrequently available.

`bayesnec` takes a short formula naming a concentration-response equation, and
combined with the data input constructs the full non-linear `brms` formula,
generates priors and initial values, chooses a statistical distribution
appropriate to the response, fits the model, and extracts toxicity estimates from
it.

The idea was to may non-linear Bayesian concentration-response modelling accessible to 
ecotoxicologists without specialist statistical expertise. 

## Package history

The package began as `jagsNEC` [@Fisher2020], an implementation of the *NEC*
model of @Fox2010 using JAGS, which is one fo the other widely used Bayesian packages available in R,
based of one of the earliest Bayesian software pacakges `winbugs`. 

`bayesnec` was rewritten based on `brms` and Stan, largely because of `brms` extensive
set of helper and other support functions for working with Bayesian model output.

`bayesnec` has also been
generalised well beyond the original: the supported response distributions now
include the beta-binomial alongside the Gaussian, Poisson, binomial, gamma,
negative binomial and beta of the original. More recently, the dev version has also had some of the `brms` hurdle families. 

Any `brms` family can in
principle be added.

The model set was also extended. Alongside the threshold models that estimate an
*NEC* directly,  `bayesnec` includes smooth concentration-response equations of the
kind used in most frequentist packages such as `drc` [@Ritz2015] — three-, four- and five- parameter
logistic and Weibull curves among them — which model the response as a smooth
function of concentration rather than with a breakpoint. Module 3 sets out the
full set and what each one supports.

## Versions

The versions in use when a model is fitted should always be recorded as part of the result. 
This is because sometimes the
default settings used and the behaviour of estimators can change between releases. 

Recording them costs one line:

```{r}
#| label: versions
#| comment: ""
for (p in c("bayesnec", "brms", "cmdstanr", "rstan", "posterior")) {
  cat(sprintf("%-10s %s\n", p, as.character(packageVersion(p))))
}
cat(sprintf("%-10s %s\n", "R", getRversion()))
```

This course is built against the development version of `bayesnec`, which is
ahead of the release on CRAN. The setup instructions refer the version to install.

## Documentation and support

The `bayesnec` reference site at
[open-aims.github.io/bayesnec](https://open-aims.github.io/bayesnec/) is the
place to start.

![The `bayesnec` reference site. *Reference* documents every exported function,
*Articles* covers single-model and multi-model usage, the model set, priors and
comparing posterior predictions, and *Changelog* records what changed between
versions.](images/bayesnec_website.png){#fig-website}

*Reference* is the authoritative description of every argument, and is worth
preferring over any transcription of it, including the ones in this course.
*Changelog* matters more than it usually does for a package under active
development: estimator behaviour and default priors have both changed between
recent versions.

The package is developed as open source software. We want to grow `bayesnec` andmake it more useful to people.
Please provide feedback and report software bugs.

Please use the [issues](https://github.com/open-aims/bayesnec/issues) page to report any problems.

We can only fix bugs or help with issues if we can reproduce them. 

Tis requires a "reprex": a self-contained reproducible example. 

This is essentially code and
data that someone else can run to see the same behaviour. 
Without a reproducible example a problem cannot
be diagnosed by anyone who does not have your dataset. 
A good resource for understanding how to build a reprex is the [reprex package](https://reprex.tidyverse.org/).
