---
title: "Fitting a single model"
subtitle: "One concentration-response model in `bayesnec`, start to finish"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(bayesnec)
options(mc.cores = 4)
set.seed(333)
```

## Background

`bayesnec` fits concentration-response curves to toxicity data and derives
no-effect concentration (*NEC*), no-significant-effect concentration (*NSEC*,
@fisherfox2023) and effect concentration (**EC~x~**) estimates from non-linear
models fitted by Bayesian Hamiltonian Monte Carlo, via `brms`
[@Burkner2017; @Burkner2018] and Stan. It is an extension of `jagsNEC`
[@Fisher2020].

This module fits one model to one dataset and works through what is done with
the result: looking inside the fitted object, plotting it, and extracting
toxicity estimates. Later modules cover the full model set, model averaging, the
choice of statistical distribution, priors, and the checks that go with each.

If you have not yet installed R, a C++ compiler, CmdStan and `bayesnec`, work
through the [software setup](0Software-setup.qmd) first. Nothing in this module
will run without them.

## The example dataset

`bayesnec` supplies a small example dataset, `nec_data`, with two columns: `x`,
the concentration, and `y`, the response.

```{r}
#| label: data
library(bayesnec)
data(nec_data)
head(nec_data)
```

The response here is a proportion bounded by zero and one, which matters for the
statistical distribution used to model it. `bayesnec` will choose a distribution
from the characteristics of the data unless told otherwise, and module 5 covers
that choice in more detail.

Supply the response as it was measured. Do not convert it to percent inhibition,
percent of control, or percent of the observed maximum before calling `bnec()`.
Dividing by a quantity estimated from the same data biases the effect
concentrations read off the result, and `bnec()` reports the practice where it
detects it. Module 5 sets out the measurement behind that and gives the
alternative, which is to choose a family that suits the response on its own
scale.

## Reproducibility

A Bayesian model is fitted by drawing samples, which module 6 describes, so
running the same code twice gives slightly different numbers unless the random
number generator is set to a known state.

Two generators are involved and both have to be fixed. `bnec()` searches for
valid starting values in R before it calls Stan, and that search draws from R's
own generator, which `set.seed()` fixes. Stan then samples using a generator of
its own, which the `seed` argument fixes. Setting one and not the other leaves
the fit irreproducible, so every fit in this course sets both.

```{r}
#| label: seed
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)
```

The estimates below will not match yours to the last decimal place unless you
use the same seed, the same number of chains and the same version of the
package. They should match to within the width of the credible intervals, and if
they do not, that is worth investigating rather than accepting.

## The `bnec` function

`bnec` is the working function of the package. It assembles the non-linear model
formula, generates priors and initial values, and calls `brms`.

Beyond the formula and the data, its arguments fall into four groups. Run
`?bnec` for the full list.

- **The model or set to fit:** `model`.
- **How toxicity estimates are computed from the fit:** `x_range`,
  `resolution`, `sig_val`.
- **How the response and predictor are identified** where the formula alone is
  not enough: `x_var`, `y_var`, `trials_var`.
- **How priors are set:** `prior`, `prior_type`.

Anything else is passed through to `brms::brm()` by `...`, so the full `brm`
argument list is available. `bayesnec` sets three of those arguments differently
from `brms`: it generates its own `family` and `prior` rather than taking the
`brms` defaults, and it raises `iter` from 2,000 to 10,000 with a
correspondingly longer warm-up, because the non-linear models here often need
it.

Two arguments are the ones to reach for when a fit is marginal:

- Raise `iter` and `warmup` where the convergence diagnostic is borderline or
  the effective sample size is low.
- Pass `control = list(adapt_delta = 0.99, max_treedepth = 12)` where the
  sampler reports divergent transitions or reports hitting the maximum tree
  depth.

Both increase the run time, and neither corrects a model that does not suit the
data.

::: {.callout-note collapse="true" title="Iterations, warm-up and retained draws"}

Compare the **retained** draws rather than `iter`, because the fraction
discarded during warm-up is a choice each package makes for itself.

| Package | Chains | `iter` | Discarded as warm-up | Retained draws |
|---|---|---|---|---|
| `bayesnec` | 4 | 10,000 | four fifths | 8,000 |
| `brms` defaults | 4 | 2,000 | one half | 4,000 |

`bayesnec` sets `warmup` to `floor(iter / 5) * 4`. A number of iterations means
nothing until the warm-up rule and the chain count are stated beside it, which
matters because two of the sampler diagnostics are counts read against the
retained draws.
:::

## The formula

A `bayesnec` formula takes the form:

```{r}
#| label: formula-shape
#| eval: false
y ~ crf(x, model = "modelname")
```

The left-hand side is the response, exactly as in `brms`. The right-hand side
uses `crf()`, a function specific to `bayesnec`, which stands for
**c**oncentration-**r**esponse **f**unction. It takes the predictor and the name
of the model to fit, and it accepts a transformation of the predictor, such as
`crf(log(x), ...)`.

Term names must match the columns of the data frame exactly, as in any R
formula.

The available models are:

```{r}
#| label: model-list
models()$all
```

These fall into named groups, which is how a set of models is requested rather
than a single one:

```{r}
#| label: model-groups
names(models())
```

Module 3 describes what each model is and which toxicity estimates can be derived from
it. Module 4 covers fitting a set of models rather than a single model.

## The scale of the predictor

Settle the scale of the predictor before fitting anything. It changes the curve
being fitted rather than only the axis the curve is drawn against, and it
generally matters more than the choice of equation.

- The predictor enters the equations directly. `bnec()` takes the concentration
  exactly as it is given and does not make this decision for you.
- Read the spacing of the **design** rather than the shape of the response.
  Concentrations near evenly spaced after a logarithmic transformation were laid
  out on a log scale, and are usually best modelled there.
- Write the transformation inside `crf()`, as
  `crf(log(x), model = "nec4param")`.
- Add any offset to the data yourself first. A control recorded as `0` arrives
  at the data check as `log(0)` and the fit stops.
- Back-transform the estimates through the `xform` argument, which `nec()`,
  `ecx()`, `nsec()` and `autoplot()` all take.
- State the scale alongside any value you report.

A dilution series makes each concentration a fixed multiple of the one below it.
On a linear axis such a design crushes the low concentrations together and
spreads the high ones out, so the equation spends its flexibility at the top of
the range, where the response is often already flat. The coefficient of
variation of the gaps separates the two cases. The ametryn series in the
`herbicide` data that ships with `bayesnec` is a three-fold dilution from 0.1 to
300 $\mu g L^{-1}$, twelve replicates at each of eight concentrations:

```{r}
#| label: spacing
spacing_cv <- function(x) sd(diff(sort(x))) / mean(diff(sort(x)))

data(herbicide)
ametryn <- subset(herbicide, herbicide == "ametryn")
conc <- unique(ametryn$concentration)

round(c(linear = spacing_cv(conc), log = spacing_cv(log(conc))), 2)
```

The coefficient of variation is 1.72 on the linear axis and 0.05 on the log
axis, which is the signature of a series designed on the log scale.

Where the predictor is transformed, `xform` takes the inverse and returns the
estimates as concentrations:

```{r}
#| label: xform-demo
#| eval: false
ecx(fit, xform = function(x) exp(x) - 1)
autoplot(fit, xform = function(x) exp(x) - 1)
```

::: {.callout-note collapse="true" title="The size of the difference between scales"}

Fitting the ametryn series on each axis in turn measures what the scale does
here. Three equations are fitted on `log(concentration)` and `nec4param` alone on
the raw concentration axis. All four use the same response and the same 96
observations, so `elpd_loo` is comparable between them.

```{r}
#| label: scale-fit
#| eval: false
set.seed(333)
m2_scale_log <- bnec(fvfm ~ crf(log(concentration),
                                model = c("nec4param", "ecx4param", "ecxll4")),
                     data = ametryn, seed = 333)
save(m2_scale_log, file = "fits/m2_scale_log.RData")

set.seed(333)
m2_scale_raw <- bnec(fvfm ~ crf(concentration, model = "nec4param"),
                     data = ametryn, seed = 333)
save(m2_scale_raw, file = "fits/m2_scale_raw.RData")
```

```{r}
#| label: scale-load
load("fits/m2_scale_log.RData")
load("fits/m2_scale_raw.RData")
```

```{r}
#| label: scale-compare
#| message: false
#| warning: false
elpd <- function(fit) {
  brms::loo(pull_brmsfit(fit))$estimates["elpd_loo", "Estimate"]
}

# The largest gap between the fitted mean and the observed mean at a tested
# concentration, which is the departure a plot of the curve would show.
max_departure <- function(fit) {
  by_conc <- aggregate(
    cbind(observed, fitted) ~ concentration,
    data.frame(concentration = ametryn$concentration,
               observed = ametryn$fvfm,
               fitted = fitted(pull_brmsfit(fit))[, "Estimate"]),
    mean)
  max(abs(by_conc$observed - by_conc$fitted))
}

scale_fits <- list(nec4param_raw = m2_scale_raw,
                   nec4param_log = pull_out(m2_scale_log, "nec4param"),
                   ecx4param_log = pull_out(m2_scale_log, "ecx4param"),
                   ecxll4_log    = pull_out(m2_scale_log, "ecxll4"))

data.frame(elpd_loo = round(sapply(scale_fits, elpd), 1),
           max_departure = round(sapply(scale_fits, max_departure), 3))
```

Fitting `nec4param` on the log axis rather than the raw axis raises `elpd_loo`
by 23.7. The three equations fitted on the log axis span 4.3 between them. On these data the
scale separates the fits by about five times what the choice of equation does.

The group means are reproduced on both axes. The largest departure of the fitted
mean from the observed mean at a tested concentration is 0.041 on the raw axis
and 0.032 on the log axis, on a response spanning 0.7, so the difference is not
one a plot of the curve would show. `elpd_loo` is what reports it.
:::

::: {.callout-note collapse="true" title="Zero on the predictor axis"}

The transformation is evaluated before `bayesnec` inspects the predictor, which
is why the fit stops with a report that the predictor is not finite rather than
with anything naming the transformation.

`bayesnec` does not correct a zero on the predictor axis. A concentration of
zero is an ordinary control, and no response distribution constrains the values
a predictor may take. Choose the offset yourself and add it to the data first,
as `log(x + 1)` or a smaller constant suited to the series.
:::

::: {.callout-note collapse="true" title="Reporting a value on the fitted scale"}

A value reported on the fitted scale reads as a concentration and is not one.
Reporting a logged value as a concentration understates the threshold by a
factor of e for every unit on the logged scale, and only a negative logarithm
announces the error.
:::

`nec_data` is left on the linear scale throughout this course, and deliberately
so. It is simulated as 100 random draws rather than as a designed series, and
the true curve is defined as a function of linear `x`, so logging it would
distort the generating scale rather than reveal it. The difference between this
example and a real dilution series is the design rather than the package.

## Calling `bnec`

A formula and a data frame are the minimum required to fit a `bayesnec` model,
and every other argument has a default. Understand what those defaults are and
whether they suit your data; for this example they do.

```{r}
#| label: fit
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)

save(bnec_fit, file = "fits/m2_bnec_fit.RData")
```

```{r}
#| label: load-fit
load("fits/m2_bnec_fit.RData")
```

Three things happen when that call runs.

- **Compilation.** Stan models are written in the Stan language, translated to
  C++ and compiled to a shared object, which R then loads and runs. This is why
  the first fit of a session is slower than later ones, and why a compiler is
  needed at all.
- **Sampling.** How long this takes depends on the machine, the number of chains
  and the number of iterations. The fit above took a little over two minutes on
  four cores.
- **Reporting.** `bayesnec` names the model that was fitted and the statistical
  distribution used, here a `nec3param` model with a beta distribution.

Set `mc.cores` to at least the number of chains before fitting, so that the
chains run at the same time rather than in turn:

```{r}
#| label: cores
#| eval: false
options(mc.cores = 4)
```

Warnings may also appear when you run the fit yourself, which the page does
not, since it loads a saved object. The two this fit produces concern the Pareto
*k* diagnostic and `p_waic`, both of which relate to the leave-one-out
cross-validation used to compute model weights. Weights are irrelevant when a
single model has been fitted, and module 4 returns to them when they are not.

::: {.callout-note collapse="true" title="Chains in series and in parallel"}

A model is fitted by running several chains, and those chains are independent of
one another until the results are combined. Run in **series** they wait their
turn and the fit takes as long as all of them added together. Run in **parallel**
they run at the same time on different processor cores, and the fit takes about
as long as the slowest one.

More cores than chains gives no benefit here, because there is nothing else to
run at the same time. A second level of parallelism applies once a *set* of
models is being fitted rather than one, since the models can run alongside each
other as well as the chains within them. Module 4 covers it.
:::

## Saving the fit

Fit once, save, and load the saved object everywhere afterwards. Sampling is
what takes the minutes and it never needs repeating.

```{r}
#| label: save-demo
#| eval: false
saveRDS(bnec_fit, "nec3param_fit.rds")
bnec_fit <- readRDS("nec3param_fit.rds")
```

Every module in this course is built that way, and so is a real analysis:

- The fitting chunk is shown rather than run, and it ends in a `save()`.
- The chunk that runs is the `load()`, which restores the object under the name
  it was saved with. `bnec_fit` then exists exactly as though it had just been
  fitted.
- Fit in one script, and plot, summarise and extract estimates in another. A fit
  that is never saved is refitted every time a figure changes.

The workshop distributes these saved objects, so the modules load in seconds
rather than sampling for minutes. The [software setup](0Software-setup.qmd)
covers retrieving them. Fitting the models yourself works and is slower, which
is worth doing at least once to see how long it takes.

::: {.callout-note collapse="true" title="Where to put the file"}

This is one of the smallest fits in the course and the object is
`r format(object.size(bnec_fit), units = "Mb")`, because it stores every
posterior draw for every parameter. A model-averaged set of all
`r length(models()$all)` models is very much larger, and a fit to real data with
a grouping variable larger again.

A folder synchronised by OneDrive or a similar service is a poor choice: the
upload is slow, and on some configurations it interferes with writing the file
at all.
:::

## The summary

```{r}
#| label: summary
summary(bnec_fit)
```

Five things are printed, in the order they appear.

- **The class and the model.** A `bayesnecfit` containing `nec3param`.
  `bayesnec` returns a `bayesnecfit` when one model is fitted and a
  `bayesmanecfit` when a set is fitted.
- **The family and link.** The beta distribution with an **identity link**.
  `bayesnec` uses the identity link for every family, which is not the `brms`
  default.
- **The `brms` formula that was fitted.** For `nec3param` this is a constant
  `top` until the concentration reaches `nec`, followed by exponential decay,
  with the switch implemented by `step()`. Module 3 sets out the equations.
- **The coefficients, with *Rhat* and effective sample size.** *Rhat* is a
  convergence diagnostic and should be close to 1. All three parameters here are
  at 1.00 and the effective sample sizes are in the thousands, both of which
  indicate the sampler explored the posterior properly.
- **The *NEC* estimate and the Bayesian R².** The estimate is a median with a
  95% credible interval. The R² here is high, as you would expect of a simulated
  example where the curve is well-determined; real data rarely look this tidy.

::: {.callout-note collapse="true" title="The identity link"}

The link will mean something only to a reader familiar with generalised linear
models, which are usually parameterised for linear modelling. A beta response in
`brms` defaults to the logit link.

`bayesnec` overrides that for every family, so that the `top`, `bot` and `nec`
parameters stay on the scale of the response and remain directly interpretable.
It matters if you ever compare a `bayesnec` fit against a `brm` model you have
written yourself: unless you pass the link explicitly, the two are different
models.
:::

## The fitting record

`bnec()` reports what it did as it goes, through messages. A knitted document or
a call wrapped in `suppressMessages()` keeps none of them, so the same
information is stored on the fitted object and read back with `bnec_record()`.

```{r}
#| label: record
rec <- bnec_record(bnec_fit)
names(rec)
```

Two of those elements are the ones to read after a fit.

`excluded` names every equation dropped before fitting, with the reason. An
equation is excluded where it is invalid for the family or the predictor.
`requested` divides exactly between `attempted` and `excluded`, and an equation
that was attempted and then failed to fit stays in `attempted`, where module 4
covers how to find it.

```{r}
#| label: record-excluded
rec$excluded
```

Nothing was excluded here, because one model was requested and it is valid for a
beta response.

`substitutions` names every value `bayesnec` altered because the family cannot
represent it as recorded. A beta distribution is defined on the open interval
between 0 and 1, so a response of exactly 0 or exactly 1 lies outside it;
`bayesnec` replaces a zero with one tenth of the smallest non-zero value and
reduces a one by 0.001.

```{r}
#| label: record-subs
rec$substitutions
```

`NULL` means the response was fitted as recorded. Where it is not `NULL`, the
analysis was run on the substituted values and a methods section has to say so.
Module 5 returns to this, because a series with several concentrations sitting
at a boundary is being modelled through the substitution rather than as
measured.

## Checking the fit

Two things have to hold before an estimate is taken from a fit, and they are
different things: that the sampler explored the posterior, and that the fitted
model reproduces the data. A model can converge perfectly onto a description of
the data that is wrong.

`bayesnec` answers each with one function, and both apply to a single fit and to
a model set alike:

| Function | Question it answers | Screens on |
|---|---|---|
| `check_sampling()` | did the sampler explore the posterior | R-hat, effective sample size, divergent transitions |
| `check_fit()` | does the model reproduce the data | observed against simulated mean and spread, within groups of the predictor |

```{r}
#| label: check-sampling
check_sampling(bnec_fit)
```

`failed` is `FALSE`, so this fit clears all three sampler cutoffs. Module 4 sets
out what each of the three measures and what to do when one is missed.

```{r}
#| label: check-fit
#| warning: false
#| fig-height: 4.5
cf <- check_fit(bnec_fit)
plot(cf)
```

`check_fit()` returns a table of one row per group of the predictor and plots
the same comparison. Each panel shows the observed statistic against the 95%
span of what the model simulates, with location and scale drawn separately
because they fail independently: a model can reproduce every group's mean while
simulating far too much spread. A point outside its bar is a group the model
does not reproduce, and no point falls outside its bar here.

The message `check_fit()` prints flags one group on a stricter criterion than
the bars use, so the two can disagree. Read it as a prompt to look at that
group rather than as a verdict on the fit.

Module 4 sets out the table `check_fit()` returns, what the control row
contributes to an *NSEC*, and how both checks behave across a set of
twenty-three equations. Module 7 shows what they contribute to a written
analysis, on real data. Leave the detail until then; for a first fit, run both
and read `failed`.

The visual form of the sampling check is worth seeing once. Extract the
underlying `brms` model and plot it, which gives the marginal posterior of each
parameter beside its trace:

```{r}
#| label: chains
#| fig-height: 6
brm_fit <- pull_brmsfit(bnec_fit)
plot(brm_fit)
```

Read the traces from left to right. Four chains that overlap, with no drift and
no sustained excursion, indicate the chains have converged on the same
distribution. Chains that separate into bands, or wander, indicate they have
not, and the estimates should not be used. Reading plots does not scale beyond a
handful of parameters, which is why `check_sampling()` exists.

## Inside the fitted object

A `bayesnecfit` is a list holding the `brms` fit and a set of derived quantities.

```{r}
#| label: names
names(bnec_fit)
```

`pred_vals` holds fitted values across the range of the predictor, which is what
the plotting methods use and what you would use for a custom figure:

```{r}
#| label: predvals
head(bnec_fit$pred_vals$data)
```

Every fitted model holds a posterior for the no-effect estimate, in
`ne_posterior`, with `ne_type` recording what kind of estimate it is. For models
in the `nec` group this is the posterior of the `nec` parameter itself. For
models in the `ecx` group, which have no threshold parameter, it is the
posterior of the *NSEC* [@fisherfox2023]. Module 3 explains why those are
different things.

```{r}
#| label: ne-posterior
#| fig-height: 4
hist(bnec_fit$ne_posterior,
     main = "", xlab = "No-effect concentration", col = "grey80", border = "white")
```

## Plotting

`plot()` extends the base method and `autoplot()` extends the `ggplot2` method.
Both take arguments controlling what is drawn; see `?autoplot.bayesnecfit`.

```{r}
#| label: autoplot
#| fig-height: 4.5
autoplot(bnec_fit) +
  ggplot2::xlab("Concentration")
```

The result is a `ggplot` object, so it composes with `+` in the usual way, and
most `ggplot2` functions will work on it.

By default the plot is annotated with the fitted no-effect estimate and its 95%
credible interval, and labelled with the model that was fitted. The annotation
reads NEC, because for a `nec3param` model the no-effect estimate is the *NEC*.
Module 3 covers what that label becomes for the other equations, and module 4
covers how these estimates are combined across a model set.

## Toxicity estimates

Three functions extract estimates from a fitted model: `ecx()`, `nec()` and
`nsec()`.

```{r}
#| label: estimates
ecx(bnec_fit)
nec(bnec_fit)
nsec(bnec_fit)
```

Each returns a median and a 95% probability interval by default, and each
records how it was computed in the attributes of the returned object: `ecx()`
returns an **EC10** unless told otherwise, and `nsec()` uses a significance
value of 0.01.

For this dataset the three estimates are close together, which happens when the
curve declines sharply from a well-defined threshold. They are not
interchangeable, and on other data they can differ. Module 3 covers what
each of them measures and when each is appropriate.

## The parameters of the curve

`summary()` reports the toxicity estimate and the fit statistics, and for a
single model the parameters are reachable through the underlying `brmsfit`.
`curve_params()` returns them directly, and returns them for a model-averaged set
as well, which nothing else does:

```{r}
#| label: curve-params
curve_params(bnec_fit)
```

These are the quantities the toxicity estimates are read from, and they belong in
a methods section beside them. `top` is the control level the curve is referenced
to, `beta` is the decay rate, and `nec` is the threshold. A four-parameter
equation would add `bot`, the lower asymptote.

`xform` applies to `nec` and `ec50` and to no other parameter, because those two
are measured on the predictor axis while `top`, `bot` and `beta` are levels of
the response or shape parameters, on which a transformation of the predictor has
no meaning.

## References

::: {#refs}
:::
