---
title: "Fitting a single model"
subtitle: "One concentration-response model in `bayesnec`, start to finish"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(bayesnec)
options(mc.cores = 4)
```

## Background

`bayesnec` fits concentration-response curves to toxicity data and derives
no-effect concentration (*NEC*), no-significant-effect concentration (*NSEC*,
@fisherfox2023) and effect concentration (**EC~x~**) estimates from non-linear
models fitted by Bayesian Hamiltonian Monte Carlo, via `brms`
[@Burkner2017; @Burkner2018] and Stan. It is an extension of `jagsNEC`
[@Fisher2020].

This module fits one model to one dataset and works through everything that can
be done with the result: checking that it converged, looking inside the fitted
object, plotting it, and extracting toxicity estimates. Later modules cover the
full model set, model averaging, the choice of statistical distribution, and
priors.

If you have not yet installed R, a C++ compiler, CmdStan and `bayesnec`, work
through the [software setup](0Software-setup.qmd) first. Nothing in this module
will run without them.

## The example dataset

`bayesnec` supplies a small example dataset, `nec_data`, with two columns: `x`,
the concentration, and `y`, the response.

```{r}
#| label: data
library(bayesnec)
data(nec_data)
head(nec_data)
```

The response here is a proportion bounded by zero and one, which matters for the
statistical distribution used to model it. `bayesnec` will choose a distribution
from the characteristics of the data unless told otherwise, and module 5 covers
that choice in full.

## Reproducibility

A Bayesian model is fitted by drawing samples, so running the same code twice
gives slightly different numbers unless the random number generator is set to a
known state. `set.seed()` does that, and every fit in this course is preceded by
one.

```{r}
#| label: seed
#| eval: false
set.seed(333)
```

The estimates below will not match yours to the last decimal place unless you
use the same seed, the same number of chains and the same version of the
package. They should match to within the width of the credible intervals, and if
they do not, that is worth investigating rather than accepting.

## The `bnec` function

`bnec` is the working function of the package. It assembles the non-linear model
formula, generates priors and initial values, and calls `brms`. It is designed
for an R user who is comfortable with model-fitting syntax but is not a
specialist in non-linear or Bayesian modelling.

Run `?bnec` for the full argument list. Beyond the formula and the data, the
arguments fall into four groups: the model or model set to fit (`model`); how
toxicity estimates are computed from the fit (`x_range`, `resolution`,
`sig_val`); how the response and predictor are identified when the formula alone
is not enough (`x_var`, `y_var`, `trials_var`); and how priors are set (`prior`,
`prior_type`).

Anything else is passed through to `brms::brm()` by `...`, so the full `brm`
argument list is available. Three `brm` arguments are worth knowing about
because `bayesnec` sets them differently from `brms`. It generates its own
`family` and `prior` values rather than using the `brms` defaults, and it raises
the number of iterations from 2,000 to 10,000, with a correspondingly longer
warm-up, because the non-linear models here often need it.

## The formula

A `bayesnec` formula takes the form:

```{r}
#| label: formula-shape
#| eval: false
y ~ crf(x, model = "modelname")
```

The left-hand side is the response, exactly as in `brms`. The right-hand side
uses `crf()`, a function specific to `bayesnec`, which stands for
**c**oncentration-**r**esponse **f**unction. It takes the predictor and the name
of the model to fit, and it accepts a transformation of the predictor, such as
`crf(log(x), ...)`.

Term names must match the columns of the data frame exactly, as in any R
formula.

The available models are:

```{r}
#| label: model-list
models()$all
```

These fall into named groups, which is how a set of models is requested rather
than a single one:

```{r}
#| label: model-groups
names(models())
```

Module 3 describes what each model does and which toxicity estimates follow from
it. Module 4 covers fitting a set rather than a single model.

## Calling `bnec`

Supplying a formula and a data frame is enough; everything else takes a default.

```{r}
#| label: fit
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data)
```

The first thing that happens is compilation. Stan models are written in the Stan
language, translated to C++ and compiled to a shared object, which R then loads
and runs. This is why the first fit of a session is slower than later ones, and
why a compiler is needed at all.

Sampling follows. How long the whole thing takes depends on the machine, the
number of chains and the number of iterations. The fit above took a little over
two minutes on four cores.

A model is fitted by running several chains, and those chains are independent of
one another until the results are combined. Run in **series** they wait their
turn and the fit takes as long as all of them added together; run in
**parallel** they run at the same time on different processor cores and the fit
takes about as long as the slowest one. Setting `mc.cores` is what allows the
second, so use at least as many cores as you have chains:

```{r}
#| label: cores
#| eval: false
options(mc.cores = 4)
```

More cores than chains gives no benefit here, because there is nothing else to
run at the same time. There is a second level of parallelism once a *set* of
models is being fitted rather than one — the models can run alongside each other
as well as the chains within them — and module 4 covers it.

When the fit completes, `bayesnec` reports which model was fitted and which
statistical distribution was used. Here it is a `nec3param` model with a beta
distribution.

Warnings may also appear. The two above concern the Pareto *k* diagnostic and
`p_waic`, both of which relate to the leave-one-out cross-validation used to
compute model weights. Weights are irrelevant when a single model has been
fitted, and module 4 returns to them when they are not.

## Saving the fit

A fitted model is expensive to produce and cheap to store, so save it as soon as
it finishes.

```{r}
#| label: save-demo
#| eval: false
saveRDS(bnec_fit, "nec3param_fit.rds")
bnec_fit <- readRDS("nec3param_fit.rds")
```

Be deliberate about where. This is one of the smallest fits in the course and the
object is `r format(object.size(bnec_fit), units = "Mb")`, because it stores
every posterior draw for every parameter. A model-averaged set of all
`r length(models()$all)` models is very much larger, and a fit to real data with
a grouping variable larger again. A folder synchronised by
OneDrive or a similar service is a poor choice: the upload is slow, and on some
configurations it interferes with writing the file at all.

## The summary

```{r}
#| label: summary
summary(bnec_fit)
```

There is a lot here, and each part is worth reading.

The object is of class `bayesnecfit`, containing the `nec3param` model.
`bayesnec` returns a `bayesnecfit` when one model is fitted, and a
`bayesmanecfit` when a set is fitted.

The family is the beta distribution with an **identity link**. This is
deliberate and it is not the `brms` default, which for a beta response is the
logit. `bayesnec` uses the identity link for every family, so that the `top`,
`bot` and `nec` parameters stay on the scale of the response and remain directly
interpretable. It matters if you ever compare a `bayesnec` fit against a `brm`
model you have written yourself: unless you pass the link explicitly, the two
are different models.

The `brms` formula that was actually fitted is printed. For `nec3param` it is a
constant `top` until the concentration reaches `nec`, followed by exponential
decay, with the switch implemented by `step()`. Module 3 sets out the equations.

*Rhat* is a convergence diagnostic and should be close to 1. All three
parameters here are at 1.00, and the effective sample sizes are in the
thousands, both of which indicate the sampler explored the posterior properly.

The *NEC* estimate is given below the coefficients, as a median with a 95%
credible interval, and the Bayesian R² gives an indication of fit. The value here
is high, as you would expect of a simulated example where the curve is
well-determined; real data rarely look this tidy.

## Checking the fit

`brms` offers a range of diagnostics, and the `brms` documentation covers them
far better than this course can. Extract the underlying model with
`pull_brmsfit()`:

```{r}
#| label: pull
brm_fit <- pull_brmsfit(bnec_fit)
```

Plotting it gives the marginal posterior of each parameter beside its trace:

```{r}
#| label: chains
#| fig-height: 6
plot(brm_fit)
```

Read the traces from left to right. Four chains that overlap, with no drift and
no sustained excursion, indicate the chains have converged on the same
distribution. Chains that separate into bands, or wander, indicate they have
not, and the estimates should not be used.

## Inside the fitted object

A `bayesnecfit` is a list holding the `brms` fit and a set of derived quantities.

```{r}
#| label: names
names(bnec_fit)
```

`pred_vals` holds fitted values across the range of the predictor, which is what
the plotting methods use and what you would use for a custom figure:

```{r}
#| label: predvals
head(bnec_fit$pred_vals$data)
```

Every fitted model carries a posterior for the no-effect estimate, in
`ne_posterior`, with `ne_type` recording what kind of estimate it is. For models
in the `nec` group this is the posterior of the `nec` parameter itself. For
models in the `ecx` group, which have no threshold parameter, it is the
posterior of the *NSEC* [@fisherfox2023]. Module 3 explains why those are
different things.

```{r}
#| label: ne-posterior
#| fig-height: 4
hist(bnec_fit$ne_posterior,
     main = "", xlab = "No-effect concentration", col = "grey80", border = "white")
```

## Plotting

`plot()` extends the base method and `autoplot()` extends the `ggplot2` method.
Both take arguments controlling what is drawn; see `?autoplot.bayesnecfit`.

```{r}
#| label: autoplot
#| fig-height: 4.5
autoplot(bnec_fit) +
  ggplot2::xlab("Concentration")
```

The result is a `ggplot` object, so it composes with `+` in the usual way, and
most `ggplot2` functions will work on it.

By default the plot is annotated with the fitted no-effect estimate and its 95%
credible interval, and labelled with the model that was fitted. The annotation
is labelled N(S)EC, because the same plotting code serves both model groups. For
a `nec3param` fit the quantity is an *NEC*, and it is worth relabelling it as
such in anything you present.

## Toxicity estimates

Three functions extract estimates from a fitted model: `ecx()`, `nec()` and
`nsec()`.

```{r}
#| label: estimates
ecx(bnec_fit)
nec(bnec_fit)
nsec(bnec_fit)
```

Each returns a median and a 95% probability interval by default, and each
records how it was computed in the attributes of the returned object: `ecx()`
returns an **EC10** unless told otherwise, and `nsec()` uses a significance
value of 0.01.

For this dataset the three estimates are close together, which happens when the
curve declines sharply from a well-defined threshold. They are not
interchangeable, and on other data they separate widely. Module 3 covers what
each of them measures and when each is appropriate.
