---
title: "Fitting a single model"
subtitle: "One concentration-response model in `bayesnec`, start to finish"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(bayesnec)
options(mc.cores = 4)
set.seed(333)
```

## Background

`bayesnec` fits concentration-response curves to toxicity data and derives
no-effect concentration (*NEC*), no-significant-effect concentration (*NSEC*,
@fisherfox2023) and effect concentration (**EC~x~**) estimates from non-linear
models fitted by Bayesian Hamiltonian Monte Carlo, via `brms`
[@Burkner2017; @Burkner2018] and Stan. It is an extension of `jagsNEC`
[@Fisher2020].

This module fits one model to one dataset and works through everything that can
be done with the result: checking that it converged, looking inside the fitted
object, plotting it, and extracting toxicity estimates. Later modules cover the
full model set, model averaging, the choice of statistical distribution, and
priors.

If you have not yet installed R, a C++ compiler, CmdStan and `bayesnec`, work
through the [software setup](0Software-setup.qmd) first. Nothing in this module
will run without them.

## The example dataset

`bayesnec` supplies a small example dataset, `nec_data`, with two columns: `x`,
the concentration, and `y`, the response.

```{r}
#| label: data
library(bayesnec)
data(nec_data)
head(nec_data)
```

The response here is a proportion bounded by zero and one, which matters for the
statistical distribution used to model it. `bayesnec` will choose a distribution
from the characteristics of the data unless told otherwise, and module 5 covers
that choice in more detail.

## Preparing the response

Supply the response as it was measured. Do not convert it to percent inhibition,
percent of control, or percent of the observed maximum before calling `bnec()`.

Normalising to the control is conventional in ecotoxicology, usually as
$y^0_i = 1 - y_i / \bar{y}_{con}$, where $\bar{y}_{con}$ is the mean of three to
six control replicates. @Ritz2026 measured what it does. Every observation is
divided by the *same random quantity*, so the normalised values are correlated
and the sampling variability of the divisor is discarded. Because the expected
value of a reciprocal is not the reciprocal of an expected value, the inhibition
trend is biased downwards and the effect concentrations read off it are biased
upwards.

The size of that error, from their simulation of an ED10 with six control
replicates: normalising gave a bias of 6.8% and a coefficient of variation of
26.4%, against 2.1% and 12.7% for the same quantity estimated from the raw
response, and nominal 95% intervals covered at 90% rather than 95%. With a single
control replicate the coverage of the normalised analysis fell to 57%.

Nothing is lost by not normalising. The concentration at which inhibition rises
by $x$ per cent is the concentration at which the response falls by $x$ per cent,
so the two analyses answer the same question. The second is what `ecx()` returns
by default, as a decline relative to the *fitted* control rather than to an
observed control mean, and because it is evaluated separately within every
posterior draw the uncertainty about the control level propagates into the
credible interval instead of being thrown away.

Dividing by the observed maximum instead of the control mean is worse, on three
counts. An extreme order statistic is more variable and more biased than a mean
of three to six values. The divisor then depends on every treatment rather than
on the controls alone, so treated groups partly determine their own denominator.
And it forces exactly one observation to exactly 1, which is outside the open
interval a beta distribution is defined on, so `bnec()` has to shift it off the
boundary.

`bnec()` looks for both practices and reports them. Dividing by the observed
maximum leaves a maximum of exactly 1 attained by exactly one observation, and
dividing by the control mean leaves the control observations averaging to exactly
1. The first check is suppressed where the response lies on a rational grid,
because a genuine count-derived proportion such as 19 of 20 surviving routinely
produces a single observation at 1 and is not evidence of anything.

Where a divisor is unavoidable, as it is for a response above 1 that has to be
modelled with a beta distribution, it must be a constant fixed in advance of the
analysis: a physiological or design ceiling, or a value from accumulated
historical controls. @Ritz2026 are explicit that the problem is dividing by
something *random* rather than dividing as such. An **EC~x~** is unaffected by
the choice of a constant divisor, because it is a relative decline from the
fitted `top`, and the same holds for any linear rescaling of the response
[@Weimer2012].

Many proportional responses need no divisor at all. The maximum quantum yield
used in module 7 is a genuine physical proportion, as is a survival fraction or a
bleached-area fraction, and these are modelled as they stand.

### The reason a divisor gets reached for

Dividing by the observed maximum is often done for a reason that has nothing to
do with normalisation. A response on an arbitrary scale, such as a luminescence
reading or a cell count, is divided to put it between 0 and 1 so that a beta
distribution can be used, because the beta accommodates variability that changes
with the mean and a Gaussian does not.

That reasoning identifies a real problem and answers it indirectly. The problem
is that replicate variability at the high concentrations is not the same as at
the control. The direct answer is to keep the response on its own scale,
choose the family that suits it, and model the variability explicitly where it
changes. Module 5 covers both halves: the choice of family, and the `disp` term
for a dispersion that varies across the series.

Doing it that way keeps `top`, `bot` and `nec` in the units the experiment was
measured in, avoids the bias @Ritz2026 measured, and states the change in
variability as something estimated from the data rather than absorbed by a
transformation. It also avoids forcing an observation onto the boundary of the
beta's support, which division by a maximum does by construction.

## Reproducibility

A Bayesian model is fitted by drawing samples (which I will explain in more detail in Module 6), 
so running the same code twice
gives slightly different numbers unless the random number generator is set to a
known state.

Two generators are involved and both have to be fixed. `bnec()` searches for
valid starting values in R before it calls Stan, and that search draws from R's
own generator, which `set.seed()` fixes. Stan then samples using a generator of
its own, which the `seed` argument fixes. Setting one and not the other leaves
the fit irreproducible, so every fit in this course sets both.

```{r}
#| label: seed
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)
```

The estimates below will not match yours to the last decimal place unless you
use the same seed, the same number of chains and the same version of the
package. They should match to within the width of the credible intervals, and if
they do not, that is worth investigating rather than accepting.

## The `bnec` function

`bnec` is the working function of the package. It assembles the non-linear model
formula, generates priors and initial values, and calls `brms`. It is designed
for an R user who is comfortable with model-fitting syntax but is not a
specialist in non-linear or Bayesian modelling.

Run `?bnec` for the full argument list. Beyond the formula and the data, the
arguments fall into four groups: the model or model set to fit (`model`); how
toxicity estimates are computed from the fit (`x_range`, `resolution`,
`sig_val`); how the response and predictor are identified when the formula alone
is not enough (`x_var`, `y_var`, `trials_var`); and how priors are set (`prior`,
`prior_type`).

Anything else is passed through to `brms::brm()` by `...`, so the full `brm`
argument list is available. Three `brm` arguments are worth knowing about
because `bayesnec` sets them differently from `brms`. It generates its own
`family` and `prior` values rather than using the `brms` defaults, and it raises
the number of iterations from 2,000 to 10,000, with a correspondingly longer
warm-up, because the non-linear models here often need it.

Compare the **retained** draws rather than `iter`, because the fraction discarded
during warm-up is a choice each package makes for itself. `bayesnec` fits four
chains at `iter = 1e4` and sets `warmup` to `floor(iter / 5) * 4`, so it discards
four fifths and retains 8,000 draws. `brms` discards half, so its own defaults of
`iter = 2000` over four chains retain 4,000. A number of iterations means nothing
until the warm-up rule and the chain count are stated beside it, which matters
here because two of the diagnostics in the next section are counts read against
the retained draws.

Two arguments are the ones to reach for when a fit is marginal. Raise `iter` and
`warmup` where the convergence diagnostic is borderline or the effective sample
size is low. Pass `control = list(adapt_delta = 0.99, max_treedepth = 12)` where
the sampler reports divergent transitions or reports hitting the maximum tree
depth. Both increase the run time, and neither corrects a model that does not
suit the data.

## The formula

A `bayesnec` formula takes the form:

```{r}
#| label: formula-shape
#| eval: false
y ~ crf(x, model = "modelname")
```

The left-hand side is the response, exactly as in `brms`. The right-hand side
uses `crf()`, a function specific to `bayesnec`, which stands for
**c**oncentration-**r**esponse **f**unction. It takes the predictor and the name
of the model to fit, and it accepts a transformation of the predictor, such as
`crf(log(x), ...)`.

Term names must match the columns of the data frame exactly, as in any R
formula.

The available models are:

```{r}
#| label: model-list
models()$all
```

These fall into named groups, which is how a set of models is requested rather
than a single one:

```{r}
#| label: model-groups
names(models())
```

Module 3 describes what each model is and which toxicity estimates can be derived from
it. Module 4 covers fitting a set of models rather than a single model.

## The scale of the predictor

The predictor enters the equations directly, so the scale it is supplied on
changes the curve being fitted rather than only the axis it is drawn against.
`bnec()` takes the concentration exactly as it is given and does not make this
decision for you.

Settle it before fitting anything, because it generally matters more than the
choice of equation. On one algal dataset, fitting `nec4param` on `log(dose)`
rather than on the raw dose axis reduced the largest departure of the fitted mean
from the observed mean from 0.2115 to 0.0676, on a response spanning about 0.7,
while five different equations fitted on the log axis all landed within 1.2
`elpd_loo` of one another.

Concentration-response experiments are usually laid out as a dilution series, in
which each concentration is a fixed multiple of the one below it. On a linear
axis such a design crushes the low concentrations together and spreads the high
ones out, so the equation spends its flexibility at the top of the range, where
the response is often already flat.

Read the spacing of the **design** rather than the shape of the response.
Concentrations that are near evenly spaced after a logarithmic transformation
were laid out on a log scale, and are usually best modelled there:

```{r}
#| label: spacing
spacing_cv <- function(x) sd(diff(sort(x))) / mean(diff(sort(x)))

# a three-fold dilution series
conc <- c(0.01, 0.3, 1, 3, 10, 30, 100, 300)
round(c(linear = spacing_cv(conc), log = spacing_cv(log(conc))), 2)
```

The coefficient of variation of the gaps is much smaller on the log scale, which
is the signature of a series designed on that scale.

A transformation is written directly inside `crf()`, as
`crf(log(x), model = "nec4param")`. Three consequences follow.

The transformation is evaluated before `bayesnec` inspects the predictor, so a
control recorded as `0` arrives at the data check as `log(0)` and the fit stops,
reporting a predictor that is not finite. `bayesnec` does not correct a zero on
the predictor axis: a concentration of zero is an ordinary control, and no
response distribution constrains the values a predictor may take. Choose the
offset yourself and add it to the data first, as `log(x + 1)` or a smaller
constant suited to the series.

Toxicity estimates are then returned on the transformed scale, so `nec()`,
`ecx()` and `nsec()` need back-transforming through their `xform` argument, which
`autoplot()` also takes:

```{r}
#| label: xform-demo
#| eval: false
ecx(fit, xform = function(x) exp(x) - 1)
autoplot(fit, xform = function(x) exp(x) - 1)
```

A value reported on the fitted scale reads as a concentration and is not one.
Reporting a logged value as a concentration understates the threshold by a factor
of e for every unit on the logged scale, and only a negative logarithm announces
the error. State the scale alongside any value you report.

`nec_data` is left on the linear scale throughout this course, and deliberately
so. It is simulated as 100 random draws rather than as a designed series, and the
true curve is defined as a function of linear `x`, so logging it would distort the
generating scale rather than reveal it. The difference between this example and a
real dilution series is the design rather than the package.

## Calling `bnec`

Supplying a formula and a data frame is the minimum reqired to fit a `bayesnec` model; 
everything else is supplied by a default setting. You should always understand
what these default settings are and if they are adequate for your data, but for this example they are.

```{r}
#| label: fit
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)

save(bnec_fit, file = "fits/m2_bnec_fit.RData")
```

```{r}
#| label: load-fit
load("fits/m2_bnec_fit.RData")
```

The first thing that happens is compilation. Stan models are written in the Stan
language, translated to C++ and compiled to a shared object, which R then loads
and runs. This is why the first fit of a session is slower than later ones, and
why a compiler is needed at all.

Sampling follows. How long the whole thing takes depends on the machine, the
number of chains and the number of iterations. The fit above took a little over
two minutes on four cores.

A model is fitted by running several chains, and those chains are independent of
one another until the results are combined. Run in **series** they wait their
turn and the fit takes as long as all of them added together; run in
**parallel** they run at the same time on different processor cores and the fit
takes about as long as the slowest one. Setting `mc.cores` is what allows the
second, so use at least as many cores as you have chains:

```{r}
#| label: cores
#| eval: false
options(mc.cores = 4)
```

More cores than chains gives no benefit here, because there is nothing else to
run at the same time. There is a second level of parallelism once a *set* of
models is being fitted rather than one — the models can run alongside each other
as well as the chains within them — and module 4 covers it.

When the fit completes, `bayesnec` reports which model was fitted and which
statistical distribution was used. Here it is a `nec3param` model with a beta
distribution.

Warnings may also appear. The two above concern the Pareto *k* diagnostic and
`p_waic`, both of which relate to the leave-one-out cross-validation used to
compute model weights. Weights are irrelevant when a single model has been
fitted, and module 4 returns to them when they are not. 

## Saving the fit

A fitted model takes minutes of compute to produce and seconds to store, so save
it as soon as
it finishes.

```{r}
#| label: save-demo
#| eval: false
saveRDS(bnec_fit, "nec3param_fit.rds")
bnec_fit <- readRDS("nec3param_fit.rds")
```

Be deliberate about where. This is one of the smallest fits in the course and the
object is `r format(object.size(bnec_fit), units = "Mb")`, because it stores
every posterior draw for every parameter. A model-averaged set of all
`r length(models()$all)` models is very much larger, and a fit to real data with
a grouping variable larger again. A folder synchronised by
OneDrive or a similar service is a poor choice: the upload is slow, and on some
configurations it interferes with writing the file at all.

### Fit once, then load

Every module in this course follows the same three steps, and they are the steps
a real analysis follows rather than a convenience invented for teaching.

The model is fitted, and the fit is saved:

```{r}
#| label: fit-shape
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)

save(bnec_fit, file = "fits/m2_bnec_fit.RData")
```

That chunk is shown rather than run. Then the saved object is loaded, and it is
the loading that runs when you work through the page:

```{r}
#| label: load-shape
#| eval: false
load("fits/m2_bnec_fit.RData")
```

`load()` restores the object under the name it was saved with, so `bnec_fit`
exists from that point on exactly as though it had just been fitted.

This is worth adopting in your own work for the reason the paragraph above
gives: sampling is what takes the minutes, and it never needs repeating. Fit in
one script, save, and load in the script that plots, summarises and extracts
estimates. A fit that is never saved is refitted every time a figure changes.

The workshop distributes these saved objects, so the modules load in seconds
rather than sampling for minutes. The [software setup](0Software-setup.qmd)
covers retrieving them. Fitting the models yourself works and is slower, which
is worth doing at least once to see how long it takes.

## The summary

```{r}
#| label: summary
summary(bnec_fit)
```

There is a lot here, so we'll go through each part.

The object is of class `bayesnecfit`, containing the `nec3param` model.
`bayesnec` returns a `bayesnecfit` when one model is fitted, and a
`bayesmanecfit` when a set is fitted.

The family is the beta distribution with an **identity link**. This will only mean
something to those familiar with generalised linear models, as they are usually parameterised
for linear modelling. This is
deliberate and it is not the `brms` default, which for a beta response is the
logit. `bayesnec` uses the identity link for every family, so that the `top`,
`bot` and `nec` parameters stay on the scale of the response and remain directly
interpretable. It matters if you ever compare a `bayesnec` fit against a `brm`
model you have written yourself: unless you pass the link explicitly, the two
are different models.

The `brms` formula that was fitted is printed. For `nec3param` it is a
constant `top` until the concentration reaches `nec`, followed by exponential
decay, with the switch implemented by `step()`. Module 3 sets out the equations.

*Rhat* is a convergence diagnostic and should be close to 1. All three
parameters here are at 1.00, and the effective sample sizes are in the
thousands, both of which indicate the sampler explored the posterior properly.

The *NEC* estimate is given below the coefficients, as a median with a 95%
credible interval, and the Bayesian R² gives an indication of fit. The value here
is high, as you would expect of a simulated example where the curve is
well-determined; real data rarely look this tidy.

## The fitting record

`bnec()` reports what it did as it goes, through messages. A knitted document or
a call wrapped in `suppressMessages()` keeps none of them, so the same
information is stored on the fitted object and read back with `bnec_record()`.

```{r}
#| label: record
rec <- bnec_record(bnec_fit)
names(rec)
```

`requested` is the model or set you asked for, and it is divided exactly between
`attempted` and `excluded`. An equation is excluded before fitting where it is
invalid for the family or the predictor, and `excluded` gives the reason for
each. An equation that was attempted and then failed to fit stays in `attempted`,
and module 4 covers how to find it.

```{r}
#| label: record-excluded
rec$excluded
```

Nothing was excluded here, because one model was requested and it is valid for a
beta response.

`substitutions` records any value `bayesnec` altered because the family cannot
represent it as recorded. A beta distribution is defined on the open interval
between 0 and 1, so a response of exactly 0 or exactly 1 lies outside it.
`bayesnec` replaces a zero with one tenth of the smallest non-zero value, and
reduces a one by 0.001.

```{r}
#| label: record-subs
rec$substitutions
```

`NULL` means the response was fitted as recorded. Where it is not `NULL`, the
analysis was run on the substituted values and a methods section has to say so.
Module 5 returns to this, because a series with several concentrations sitting at
a boundary is being modelled through the substitution rather than as measured.

## Checking the sampler

`brms` offers a range of diagnostics, and the `brms` documentation covers them
far better than this course can. Extract the underlying model with
`pull_brmsfit()`:

```{r}
#| label: pull
brm_fit <- pull_brmsfit(bnec_fit)
```

Plotting it gives the marginal posterior of each parameter beside its trace:

```{r}
#| label: chains
#| fig-height: 6
plot(brm_fit)
```

Read the traces from left to right. Four chains that overlap, with no drift and
no sustained excursion, indicate the chains have converged on the same
distribution. Chains that separate into bands, or wander, indicate they have
not, and the estimates should not be used.

Reading plots does not scale beyond a handful of parameters, and it will not
scale at all to the model sets of module 4. `check_sampling()` puts the same
judgement on three numbers:

```{r}
#| label: check-sampling
check_sampling(bnec_fit)
```

Each is compared against a cutoff, and `failed` is `TRUE` where any of the three
is missed.

`max_rhat` is the largest R-hat over the parameters, and the cutoff is 1.01
following @Vehtari2021. R-hat compares the variation within each chain against
the variation between chains, so a value near 1 says the chains are describing
the same distribution.

`min_ess` is the smallest effective sample size, and the cutoff is 400. Draws
from a Markov chain are correlated, so 8,000 of them hold less information than
8,000 independent draws would; the effective sample size is how many independent
draws they are worth. The floor of 400 is an absolute one, below which the R-hat
and effective sample size estimates are themselves unreliable, and it corresponds
to 100 draws per chain at the four chains `bnec()` fits. Decide on `min_ess` and
read `min_ess_ratio` beside it: the ratio separates a model that cleared 400
because a great many draws were taken from one that cleared it efficiently.

`n_divergent` is the number of divergent transitions, and the cutoff is 10. A
divergent transition is recorded when the sampler's path through parameter space
goes numerically unstable, which happens where the posterior has curvature too
sharp for the step size chosen during warm-up. This one is a `bayesnec`
convention rather than a published recommendation: Stan's own guidance is that
any divergence means the sampler failed to explore the posterior, and the
allowance of ten reflects that these non-linear models commonly produce a few
near a parameter boundary.

The two counts are absolute, so the number of retained draws changes what each
one demands, and it changes them in opposite directions. A longer run makes the
effective sample size cutoff easier to clear and the divergence cutoff harder,
because at an unchanged divergence rate more draws produce more divergences. A
screen result is therefore a property of the settings as much as of the model,
and the settings belong in the write-up beside it. All three cutoffs are
arguments to `check_sampling()` and can be changed, which is a further reason to
report the values used.

## Checking the fit against the data

Everything above asks whether the sampler explored the posterior. None of it asks
whether the fitted model reproduces the data, and those are different questions:
a model can converge perfectly onto a description of the data that is wrong.

The distinction matters most for the *NSEC*, which module 3 introduces.
`nsec()` sets its reference at a quantile of the posterior of the control mean,
and the width of that posterior is the variability the fit estimated at the
control. A model simulating more variability at the control than the data show
places that reference lower, so the curve crosses it further along and the
reported concentration is higher and less protective. A model simulating less
variability returns a lower one. The direction is not known in advance, and no
convergence diagnostic reveals it.

One number for the whole fit will not detect this. Every family with a free
dispersion parameter estimates that parameter from the same residuals such a
number would summarise, so the parameter absorbs the discrepancy: the fit reports
close agreement overall while simulating the wrong spread at the concentrations
that determine the estimate. The check is therefore made within groups of the
predictor, which is what `check_fit()` does.

```{r}
#| label: check-fit
cf <- check_fit(bnec_fit)
cf
```

The predictor in `nec_data` is continuous, with a different value for every
observation, so there are no repeated concentrations to group by and
`check_fit()` cuts the range into bins and says so. A bin is not a design point.
On a real dilution series the groups are the tested concentrations and no binning
happens.

Each row compares a statistic computed from the observations in that group
against the same statistic computed from data simulated by the fitted model.
`mean_ratio` and `sd_ratio` are the observed values divided by the simulated
ones. `ppp_mean` and `ppp_sd` are posterior predictive p-values, each the
proportion of simulated datasets in which the simulated statistic was at least as
large as the observed one, so a value near 0.5 says the observed statistic sits
in the middle of what the model produces and a value near 0 or 1 says it sits in
a tail.

Read the p-values rather than the ratios, because a ratio says nothing about how
well determined it is. The `control` column flags the lowest value of the
predictor, which is the group both `nsec()` and `ecx()` read their reference
from, so that row is the one to look at first.

For this fit the control row agrees closely: a `sd_ratio` of 0.996 and a `ppp_sd`
of 0.50, meaning the observed spread at the control is what the model produces.
The assumption the *NSEC* depends on most is supported at the concentration where
it matters. Across the ten bins the means are reproduced throughout, with
`mean_ratio` between 0.98 and 1.06, and one bin returns a `ppp_mean` of 0.03,
which is what triggers the message. Ten observations is few for a group mean, and
the departure sits inside the 95% span in the plot below.

```{r}
#| label: check-fit-plot
#| fig-height: 4.5
plot(cf)
```

Each panel shows the observed statistic against the 95% span of what the model
simulates, with location and scale drawn separately because they fail
independently: a model can reproduce every group's mean while simulating far too
much spread. A point outside its bar is a group the model does not reproduce.
Every posterior predictive p-value in the table above lies inside 0.025 to 0.975,
so no point falls outside its bar in either panel, although the bin flagged by
the message sits close to the edge of one.

The message and the bars use different criteria, so they can disagree. The
message flags a p-value outside 0.05 to 0.95 and the bars are the 95% span, which
is 0.025 to 0.975. A group can fall beyond the first and inside the second. Read
the message as a prompt to look at that group rather than as a verdict on the
fit.

Module 4 applies both of these to a set of models rather than to one, and module
7 shows what they contribute to a written analysis.

## Inside the fitted object

A `bayesnecfit` is a list holding the `brms` fit and a set of derived quantities.

```{r}
#| label: names
names(bnec_fit)
```

`pred_vals` holds fitted values across the range of the predictor, which is what
the plotting methods use and what you would use for a custom figure:

```{r}
#| label: predvals
head(bnec_fit$pred_vals$data)
```

Every fitted model holds a posterior for the no-effect estimate, in
`ne_posterior`, with `ne_type` recording what kind of estimate it is. For models
in the `nec` group this is the posterior of the `nec` parameter itself. For
models in the `ecx` group, which have no threshold parameter, it is the
posterior of the *NSEC* [@fisherfox2023]. Module 3 explains why those are
different things.

```{r}
#| label: ne-posterior
#| fig-height: 4
hist(bnec_fit$ne_posterior,
     main = "", xlab = "No-effect concentration", col = "grey80", border = "white")
```

## Plotting

`plot()` extends the base method and `autoplot()` extends the `ggplot2` method.
Both take arguments controlling what is drawn; see `?autoplot.bayesnecfit`.

```{r}
#| label: autoplot
#| fig-height: 4.5
autoplot(bnec_fit) +
  ggplot2::xlab("Concentration")
```

The result is a `ggplot` object, so it composes with `+` in the usual way, and
most `ggplot2` functions will work on it.

By default the plot is annotated with the fitted no-effect estimate and its 95%
credible interval, and labelled with the model that was fitted. The annotation
is labelled NEC, because for a `nec3param` model the no-effect estimate is the *NEC*. We will discuss this in more 
detail in module 3, and how these are handled in model averaging in module 4.

## Toxicity estimates

Three functions extract estimates from a fitted model: `ecx()`, `nec()` and
`nsec()`.

```{r}
#| label: estimates
ecx(bnec_fit)
nec(bnec_fit)
nsec(bnec_fit)
```

Each returns a median and a 95% probability interval by default, and each
records how it was computed in the attributes of the returned object: `ecx()`
returns an **EC10** unless told otherwise, and `nsec()` uses a significance
value of 0.01.

For this dataset the three estimates are close together, which happens when the
curve declines sharply from a well-defined threshold. They are not
interchangeable, and on other data they can differ. Module 3 covers what
each of them measures and when each is appropriate.

## The parameters of the curve

`summary()` reports the toxicity estimate and the fit statistics, and for a
single model the parameters are reachable through the underlying `brmsfit`.
`curve_params()` returns them directly, and returns them for a model-averaged set
as well, which nothing else does:

```{r}
#| label: curve-params
curve_params(bnec_fit)
```

These are the quantities the toxicity estimates are read from, and they belong in
a methods section beside them. `top` is the control level the curve is referenced
to, `beta` is the decay rate, and `nec` is the threshold. A four-parameter
equation would add `bot`, the lower asymptote.

`xform` applies to `nec` and `ec50` and to no other parameter, because those two
are measured on the predictor axis while `top`, `bot` and `beta` are levels of
the response or shape parameters, on which a transformation of the predictor has
no meaning.

## References

::: {#refs}
:::
