---
title: "Fitting a single model"
subtitle: "One concentration-response model in `bayesnec`, start to finish"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(bayesnec)
options(brms.backend = "cmdstanr")
options(mc.cores = 4)
set.seed(333)
```

## Background

`bayesnec` [@fisheretal2024] fits concentration-response curves to toxicity data and derives
no-effect concentration (*NEC*), no-significant-effect concentration (*NSEC*,
@fisherfox2023) and effect concentration (*EC~x~*) estimates from non-linear
models fitted by Bayesian Hamiltonian Monte Carlo, via `brms`
[@Burkner2017; @Burkner2018] and Stan. It is an extension of `jagsNEC`
[@Fisher2020].

One model is fitted here to one dataset, and the result is then examined,
plotted and used to extract toxicity estimates. Later modules cover the full
model set, model averaging, the choice of statistical distribution, priors,
and the checks that go with each.

If you have not yet installed R, a C++ compiler, CmdStan and `bayesnec`, work
through the [software setup](0Software-setup.qmd) first. Nothing in this module
will run without them.

## The example dataset

`bayesnec` supplies a small example dataset, `nec_data`, with two columns: `x`,
the concentration, and `y`, the response.

```{r}
#| label: data
library(bayesnec)
data(nec_data)
head(nec_data)
```

The response here is a proportion bounded by zero and one, which matters for the
statistical distribution used to model it. `bayesnec` will choose a distribution
from the characteristics of the data unless told otherwise, and module 5 covers
that choice in more detail.

Supply the response as it was measured. Do not convert it to percent inhibition,
percent of control, or percent of the observed maximum before calling `bnec()`.
Dividing by a quantity estimated from the same data biases the effect
concentrations read off the result, and `bnec()` reports the practice where it
detects it. Module 5 sets out the measurement behind that and gives the
alternative, which is to choose a family that suits the response on its own
scale.

## Reproducibility

A Bayesian model is fitted by drawing samples, which module 6 describes, so
running the same code twice gives slightly different numbers unless the random
number generator is set to a known state.

Two random draws are involved. `bnec()` searches for valid starting values in R
before it calls Stan, and Stan then samples. Either `set.seed()` before the call
or the `seed` argument fixes both. `bnec()` uses `seed` to seed the search for
starting values as well as passing it to Stan, and without it Stan's seed is
drawn from R's generator, which `set.seed()` has already fixed. Every fit in
this course sets both, and `?bnec` describes the cases, such as fitting a model
set in parallel, where the two are not equivalent.

```{r}
#| label: seed
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)
```

The estimates below will not match yours to the last decimal place unless you
use the same seed, the same number of chains and the same version of the
package. They should match to within the width of the credible intervals, and if
they do not, that is worth investigating rather than accepting.

## The `bnec` function

`bnec` is the working function of the package. It assembles the non-linear model
formula, generates priors and initial values, and calls `brms`.

The formula names the equation to fit and every column the fit uses, so neither
is an argument. Beyond the formula and the data, the arguments fall into three
groups. Run `?bnec` for the full list.

- **How toxicity estimates are computed from the fit:** `x_range`,
  `resolution`, `sig_val`.
- **How priors are set:** `prior`, `prior_type`, `predictor_scale`.
- **How a model set is weighted, and when a fit that will not return is
  abandoned:** `loo_controls`, `timeout`. Module 4 covers both.

`model`, `x_var`, `y_var`, `trials_var`, `random` and `random_vars` named the
equation, the columns and the grouping terms before version 2.0. They are still
accepted and warn that the formula has been used instead.

Anything else is passed through to `brms::brm()` by `...`, so the full `brm`
argument list is available. `bayesnec` sets several of those arguments
differently from `brms`: it generates its own `family`, `prior` and `init`
rather than taking the `brms` defaults, it sets `sample_prior = "yes"` so that
the prior is sampled alongside the posterior, and it raises `iter` from 2,000 to
10,000 with a correspondingly longer warm-up, because the non-linear models here
often need it.

::: {.callout-note collapse="true" title="Iterations, warm-up and retained draws"}

Compare the retained draws rather than `iter`, because the fraction
discarded during warm-up is a choice each package makes for itself.

| Package | Chains | `iter` | Discarded as warm-up | Retained draws |
|---|---|---|---|---|
| `bayesnec` | 4 | 10,000 | four fifths | 8,000 |
| `brms` defaults | 4 | 2,000 | one half | 4,000 |

`bayesnec` sets `warmup` to `floor(iter / 5) * 4`. A number of iterations means
nothing until the warm-up rule and the chain count are stated beside it, which
matters because two of the sampler diagnostics are counts read against the
retained draws.
:::

## The formula

A `bayesnec` formula takes the form:

```{r}
#| label: formula-shape
#| eval: false
y ~ crf(x, model = "modelname")
```

The left-hand side is the response, exactly as in `brms`. The right-hand side
uses `crf()`, a function specific to `bayesnec`, which stands for
**c**oncentration-**r**esponse **f**unction. It takes the predictor and the name
of the model to fit, and it accepts a transformation of the predictor, such as
`crf(log(x), ...)`.

Term names must match the columns of the data frame exactly, as in any R
formula.

The left-hand side also takes the `brms` aterms, which name a column the
response cannot be interpreted without. They follow the response after a
vertical bar, as `y | trials(n) ~ crf(x, model = "modelname")`. `bayesnec`
supports four:

| Term | What it declares |
|---|---|
| `trials(n)` | the number of trials behind a binomial or beta-binomial count |
| `rate(n)` | the denominator a Poisson or negative binomial count was observed over |
| `cens(indicator)` | which observations are censored, with `cens(indicator, upper)` for interval censoring |
| `weights(w)` | a weight per observation |

Any other aterm is an error rather than a warning. Module 5 uses `trials()`,
`rate()` and `cens()` on the response types that need them.

The available models are:

```{r}
#| label: model-list
models()$all
```

These fall into named groups, which is how a set of models is requested rather
than a single one:

```{r}
#| label: model-groups
names(models())
```

Module 3 describes what each model is and which toxicity estimates can be derived from
it. Module 4 covers fitting a set of models rather than a single model.

## Calling `bnec`

A formula and a data frame are the minimum required to fit a `bayesnec` model,
and every other argument has a default. Understand what those defaults are and
whether they suit your data; for this example they do.

```{r}
#| label: fit
#| eval: false
set.seed(333)
bnec_fit <- bnec(y ~ crf(x, model = "nec3param"), data = nec_data, seed = 333)

save(bnec_fit, file = "vignettes/fits/m2_bnec_fit.RData")
```

```{r}
#| label: load-fit
load("vignettes/fits/m2_bnec_fit.RData")
```

Three things happen when that call runs.

- **Compilation.** Stan models are written in the Stan language, translated to
  C++ and compiled. Under `cmdstanr` the result is a program that runs outside
  R; under `rstan` it is a shared object loaded into R. This is why
  the first fit of a session is slower than later ones, and why a compiler is
  needed at all.
- **Sampling.** How long this takes depends on the machine, the number of chains
  and the number of iterations. This fit samples in under three seconds on four
  cores. Compiling it takes considerably longer, so the wait for a model this
  small is mostly the step above rather than this one.
- **Reporting.** `bayesnec` names the model that was fitted and the statistical
  distribution used, here a `nec3param` model with a beta distribution.

Set two options before fitting. `mc.cores` decides how many chains run at the
same time rather than in turn, so set it to at least the number of chains.
`brms.backend` decides which interface to Stan is used, and `brms` uses `rstan`
unless told otherwise. `rstan` compiles a model a good deal more slowly: this
fit and an `ecxll3` fit to the same data took 199 seconds together under `rstan`
and 73 under `cmdstanr` on the same four-core machine, almost all of the
difference being compilation rather than sampling.

```{r}
#| label: cores
#| eval: false
options(mc.cores = 4)
options(brms.backend = "cmdstanr")
```

Both apply for the rest of the session, so they belong at the top of a script.
Module 1 covers what `cmdstanr` and `rstan` each are.

Warnings may also appear when you run the fit yourself, which the page does
not, since it loads a saved object. The two this fit produces concern the Pareto
*k* diagnostic and `p_waic`, both of which relate to the leave-one-out
cross-validation used to compute model weights. Weights are irrelevant when a
single model has been fitted, and module 4 returns to them when they are not.

::: {.callout-note collapse="true" title="Chains in series and in parallel"}

A model is fitted by running several chains, and those chains are independent of
one another until the results are combined. Run in **series** they wait their
turn and the fit takes as long as all of them added together. Run in **parallel**
they run at the same time on different processor cores, and the fit takes about
as long as the slowest one.

More cores than chains gives no benefit here, because there is nothing else to
run at the same time. A second level of parallelism applies once a *set* of
models is being fitted rather than one, since the models can run alongside each
other as well as the chains within them. Module 4 covers it.
:::

## Saving the fit

Fit once, save, and load the saved object everywhere afterwards. Sampling is
what takes the minutes and it never needs repeating.

```{r}
#| label: save-demo
#| eval: false
saveRDS(bnec_fit, "nec3param_fit.rds")
bnec_fit <- readRDS("nec3param_fit.rds")
```

Every module in this course is built that way, and so is a real analysis:

- The fitting chunk is shown rather than run, and it ends in a `save()`.
- The chunk that runs is the `load()`, which restores the object under the name
  it was saved with. `bnec_fit` then exists exactly as though it had just been
  fitted.
- Fit in one script, and plot, summarise and extract estimates in another. A fit
  that is never saved is refitted every time a figure changes.

The workshop distributes these saved objects, so the modules load in seconds
rather than sampling for minutes. The [software setup](0Software-setup.qmd)
covers retrieving them. Fitting the models yourself works and is slower, which
is worth doing at least once to see how long it takes.

::: {.callout-note collapse="true" title="Where to put the file"}

This is one of the smallest fits in the course and the object is
`r format(object.size(bnec_fit), units = "Mb")`, because it stores every
posterior draw for every parameter. A model-averaged set of all
`r length(models()$all)` models is very much larger, and a fit to real data with
a grouping variable larger again.

A folder synchronised by OneDrive or a similar service is a poor choice: the
upload is slow, and on some configurations it interferes with writing the file
at all.
:::

## The summary

```{r}
#| label: summary
summary(bnec_fit)
```

Five things are printed, in the order they appear.

- **The class and the model.** A `bayesnecfit` containing `nec3param`.
  `bayesnec` returns a `bayesnecfit` when one model is fitted and a
  `bayesmanecfit` when a set is fitted.
- **The family and link.** The beta distribution with an **identity link**.
  `bayesnec` uses the identity link for every family unless a link is named in
  the call, where `brms` defaults to the logit link for a beta response.
- **The `brms` formula that was fitted.** For `nec3param` this is a constant
  `top` until the concentration reaches `nec`, followed by exponential decay,
  with the switch implemented by `step()`. Module 3 sets out the equations.
- **The coefficients, with *Rhat* and effective sample size.** *Rhat* is a
  convergence diagnostic and should be close to 1. All three parameters here are
  at 1.00 and the effective sample sizes are in the thousands, both of which
  indicate the sampler explored the posterior properly.
- **The *NEC* estimate and the Bayesian R².** The estimate is a median with a
  95% credible interval. The R² here is high, as you would expect of a simulated
  example where the curve is well-determined; real data rarely look this tidy.

::: {.callout-note collapse="true" title="The identity link"}

In a generalised linear model the link function relates the mean of the
response to the scale on which the model is written. A beta response in `brms`
defaults to the logit link, so a curve written for `mu` would be fitted to the
logit of the mean rather than to the mean.

`bayesnec` overrides that for every family, so that the `top`, `bot` and `nec`
parameters stay on the scale of the response and remain directly interpretable.
It matters if you ever compare a `bayesnec` fit against a `brm` model you have
written yourself: unless you pass the link explicitly, the two are different
models.
:::

## The fitting record

`bnec()` reports what it did as it goes, through messages. A knitted document or
a call wrapped in `suppressMessages()` keeps none of them, so the same
information is stored on the fitted object and read back with `bnec_record()`.

```{r}
#| label: record
rec <- bnec_record(bnec_fit)
names(rec)
```

Two of those elements are the ones to read after a fit.

`excluded` names every equation dropped before fitting, with the reason. An
equation is excluded where it is invalid for the family or the predictor.
`requested` divides exactly between `attempted` and `excluded`, and an equation
that was attempted and then failed to fit stays in `attempted`, where module 7
covers how to find it.

```{r}
#| label: record-excluded
rec$excluded
```

Nothing was excluded here, because one model was requested and it is valid for a
beta response.

`substitutions` names every value `bayesnec` altered because the family cannot
represent it as recorded. A beta distribution is defined on the open interval
between 0 and 1, so a response of exactly 0 or exactly 1 lies outside it;
`bayesnec` replaces a zero with one tenth of the smallest non-zero value and
reduces a one by 0.001.

```{r}
#| label: record-subs
rec$substitutions
```

`NULL` means the response was fitted as recorded. Where it is not `NULL`, the
analysis was run on the substituted values and a methods section has to say so.
Module 5 returns to this, because a series with several concentrations sitting
at a boundary is being modelled through the substitution rather than as
measured.

## Checking the fit

Two things have to hold before an estimate is taken from a fit, and they are
different things: that the sampler explored the posterior, and that the fitted
model reproduces the data. A model can converge perfectly onto a description of
the data that is wrong.

`bayesnec` answers each with one function, and both apply to a single fit and to
a model set alike:

| Function | Question it answers | Screens on |
|---|---|---|
| `check_sampling()` | did the sampler explore the posterior | R-hat, effective sample size, divergent transitions |
| `check_fit()` | does the model reproduce the data | observed against simulated mean and spread, within groups of the predictor |

```{r}
#| label: check-sampling
check_sampling(bnec_fit)
```

`failed` is `FALSE`, so this fit clears all three sampler cutoffs. Module 7 sets
out what each of the three measures and what to do when one is missed.

```{r}
#| label: check-fit
#| warning: false
#| fig-height: 4.5
cf <- check_fit(bnec_fit)
plot(cf)
```

`check_fit()` returns a table of one row per group of the predictor and plots
the same comparison. Because `nec_data` has no replicated concentrations, the
predictor is cut into ten bins of ten observations, with a warning. A point
outside its bar is a group the model does not reproduce, and no point falls
outside its bar here. The mean of the group from 0.726 to 0.876 sits at the top
of its bar, and printing `cf` flags it with a posterior predictive p-value of
0.027; one comparison that close to its limit is expected by chance among the
twenty in the plot.

Read the `failed` column of `check_sampling()` and the plot of `check_fit()`,
and leave the rest until the modules that need it. Module 5 uses `check_fit()`
to decide whether a statistical family reproduces the observed spread. Module 7
covers what both checks contribute to a written analysis.

`pull_brmsfit()` returns the `brms` model inside the fit, which is how any
`brms` method is reached. Plotting it gives the marginal posterior of each
parameter beside its trace:

```{r}
#| label: chains
#| fig-height: 6
brm_fit <- pull_brmsfit(bnec_fit)
plot(brm_fit)
```

Four chains that overlap, with no drift and no sustained excursion, indicate
the chains have converged on the same distribution. Reading plots does not
scale beyond a handful of parameters, which is why `check_sampling()` exists,
and module 7 reads these plots against a set of equations.

## Inside the fitted object

A `bayesnecfit` is a list holding the `brms` fit and a set of derived quantities.

```{r}
#| label: names
names(bnec_fit)
```

`pred_vals` holds fitted values across the range of the predictor, which is what
`plot()` draws and what you would use for a custom figure:

```{r}
#| label: predvals
head(bnec_fit$pred_vals$data)
```

Every fitted model holds a posterior for the no-effect estimate, in
`ne_posterior`, with `ne_type` recording what kind of estimate it is. For models
in the `nec` group this is the posterior of the `nec` parameter itself. For
models in the `ecx` group, which have no threshold parameter, it is the
posterior of the *NSEC* [@fisherfox2023]. Module 3 explains why those are
different things.

```{r}
#| label: ne-posterior
#| fig-height: 4
hist(bnec_fit$ne_posterior,
     main = "", xlab = "No-effect concentration", col = "grey80", border = "white")
```

## Plotting

`plot()` extends the base method and `autoplot()` extends the `ggplot2` method.
Both take arguments controlling what is drawn; see `?autoplot.bayesnecfit`.

```{r}
#| label: autoplot
#| fig-height: 4.5
autoplot(bnec_fit) +
  ggplot2::xlab("Concentration")
```

The result is a `ggplot` object, so it composes with `+` in the usual way, and
most `ggplot2` functions will work on it.

By default the plot is annotated with the fitted no-effect estimate and its 95%
credible interval, and labelled with the model that was fitted. The annotation
reads NEC, because for a `nec3param` model the no-effect estimate is the *NEC*.
Module 3 covers what that label becomes for the other equations, and module 4
covers how these estimates are combined across a model set.

## Toxicity estimates

Three functions extract estimates from a fitted model: `ecx()`, `nec()` and
`nsec()`.

```{r}
#| label: estimates
ecx(bnec_fit)
nec(bnec_fit)
nsec(bnec_fit)
```

Each returns a median and a 95% probability interval by default, and each
records how it was computed in the attributes of the returned object: `ecx()`
returns an *EC~10~* unless told otherwise, and `nsec()` uses a significance
value of 0.01.

For this dataset the three estimates are close together, which happens when the
curve declines sharply from a well-defined threshold. They are not
interchangeable, and on other data they can differ. Module 3 covers what
each of them measures and when each is appropriate.

## The parameters of the curve

For a single model `summary()` prints the parameters in the `brms` coefficient
table. For a model-averaged set it does not print them. `curve_params()` returns
them for both:

```{r}
#| label: curve-params
curve_params(bnec_fit)
```

These are the quantities the toxicity estimates are read from, and they belong in
a methods section beside them. `top` is the control level the curve is referenced
to, `nec` is the threshold, and `beta` is the natural logarithm of the decay
rate, since the fitted formula uses `exp(beta)`: the estimate of 0.54 is a rate
of about 1.72 per unit of concentration. A four-parameter equation would add
`bot`, the lower asymptote.

Where the predictor was transformed in the formula, the `xform` argument
returns an estimate to the measured scale. `curve_params()` applies it to `nec`
and `ec50` and to no other parameter, because those two are measured on the
predictor axis while `top`, `bot` and `beta` are levels of the response or shape
parameters. Module 7 covers the choice of scale for the predictor.

Everything above is one fit of one equation. Module 3 sets out the equations
available and what each estimate measures, module 4 fits a set of them at once,
and module 6 covers what the posterior draws behind every one of these numbers
can be used for.

## References

::: {#refs}
:::
