---
title: "Frequentist fitting with `drc`"
subtitle: "The same questions asked of a maximum-likelihood engine, and where the answers differ"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(drc)
library(bayesnec)
library(toxval)
```

## Scope

This is reference material, not part of the taught course.

`drc` [@Ritz2015] fits concentration-response models by maximum likelihood. It is
mature, widely used, and has far more in it than is shown here. The purpose of
this page is narrow: to show that most of what the course does with `bayesnec`
can also be done frequentist, and to be precise about where the two engines
answer the same question and where they do not.

For `drc` itself, better sources exist:

- Christian Ritz's [dose-response site](https://bioassay.dk/)
- the `drc` paper in *PLoS ONE*, and its supplement of worked examples
  [@Ritz2015]
- community material at [rstats4ag.org](https://rstats4ag.org/dose-response-curves.html)

The section on differences at the end is the part that is specific to this
course, and is the reason this page exists.

## Fitting

`drm()` is the fitting function. It takes a formula, the data, and a mean
function through `fct`. Here is the dataset used in module 2, fitted with a
four-parameter log-logistic curve.

```{r}
#| label: fit
data(nec_data)
model_LL4 <- drm(y ~ x, data = nec_data, fct = LL.4())
```

`LL.4()` is one of a large catalogue of mean functions. `getMeanFunctions()`
lists them.

## Reading the fit

```{r}
#| label: summary
summary(model_LL4)
```

The parameters of a four-parameter `drc` model are labelled `b`, `c`, `d` and
`e`:

| | meaning | `bayesnec` equivalent |
|---|---|---|
| `b` | slope, sometimes the hill slope | `beta` or `slope` |
| `c` | lower asymptote | `bot` |
| `d` | upper asymptote | `top` |
| `e` | inflection point | `ec50` |

Where the asymptotes span the full possible response range, `e` is the EC50.

The summary reports a standard error and a test of whether each parameter
differs from zero. That test is rarely the question of interest — a lower
asymptote indistinguishable from zero is a reason to fit a three-parameter model,
not evidence of anything about toxicity.

Residuals can be extracted and checked, which is meaningful for a continuous
response and not for binomial or count data:

```{r}
#| label: fig-resid
#| fig-height: 3.5
hist(residuals(model_LL4), main = "", xlab = "Residual", col = "grey80",
     border = "white")
shapiro.test(residuals(model_LL4))
```

## Effect concentrations

`ED()` returns effect concentrations. Two of its arguments decide what is
actually being computed, and both matter more than they appear to.

```{r}
#| label: ed-args
#| comment: ""
args(drc:::ED.drc)
```

`reference` chooses what the effect is measured **from**, and defaults to
`"control"` — the fitted response at dose zero. `type` chooses what it is
measured **to**, and defaults to `"relative"`, meaning a percentage of the
fitted response range, from the upper asymptote down to the lower one.

```{r}
#| label: ed
ED(model_LL4, c(10, 50))
```

### The word `type` does not mean the same thing in the two packages

This is the trap worth carrying away from this page.

`bayesnec::ecx()` also has a `type` argument, and its values do not correspond
to `drc`'s:

- `drc` `"relative"`, the default, is a percentage of the fitted **range**. The
  closest `bayesnec` value is `"range"`, not `"relative"`.
- `drc` `"absolute"` means an absolute **response value**, so it corresponds to
  `bayesnec`'s `"direct"`.
- `bayesnec` `"absolute"`, the default there, is a percentage decline from the
  control **to zero**. `drc` has no equivalent.

Translating `ecx(fit, type = "absolute")` into `ED(m, 10, type = "absolute")`
therefore does not ask the same question. It asks for the concentration at which
the response equals 10, which for a response bounded near 1 does not exist:

```{r}
#| label: ed-absolute
#| warning: false
try(ED(model_LL4, 10, type = "absolute", display = FALSE))
```

### When the two definitions coincide, and when they do not

They coincide exactly wherever the fitted lower limit is zero, because the
range then *is* the distance from the control to zero.

```{r}
#| label: coincide
coef(model_LL4)[["c:(Intercept)"]]
```

That is the case here, which is why this dataset gives the same EC10 either way,
and it is why three-parameter models with the lower limit fixed at zero are
unambiguous.

Give the response a genuine non-zero floor and the two part company:

```{r}
#| label: diverge
floored <- nec_data
floored$y <- 0.3 + 0.6 * floored$y
m_floor <- drm(y ~ x, data = floored, fct = LL.4())

cc <- coef(m_floor)[["c:(Intercept)"]]
dd <- coef(m_floor)[["d:(Intercept)"]]

# drc's default: 10% of the fitted range
ec10_range <- ED(m_floor, 10, display = FALSE)[1, 1]

# the same fit, asked for a 10% decline from the control, expressed as the
# equivalent percentage of the fitted range
level <- 100 * (dd - (dd - 0.10 * dd)) / (dd - cc)
ec10_control <- ED(m_floor, level, display = FALSE)[1, 1]

c(fitted_lower = cc, fitted_upper = dd,
  EC10_range_referenced = ec10_range,
  EC10_control_referenced = ec10_control)
```

The same fitted curve yields two different EC10s, differing here by about seven
per cent. Neither is wrong; they are answers to different questions. What is
wrong is reporting one and calling it the other.

The practical rule: when comparing an `ED()` result against an `ecx()` result,
establish first that both are referenced the same way, and say in the write-up
which convention was used.

## A no-effect estimate

`drc` reports no threshold parameter, so there is no *NEC* to be had from it.
The *NSEC* can be computed from a `drc` fit, because it is defined from the
fitted curve and the control variability rather than from a model parameter, and
`toxval` provides it:

```{r}
#| label: nsec
toxval::nsec(model_LL4, x_var = "x")
```

This is the one place where the frequentist route reaches a no-effect estimate
of the kind module 3 describes.

## Binomial data

A binomial response is given as successes over totals, with `weights` and
`type = "binomial"`:

```{r}
#| label: binom
data(daphnids)
daphnids.m1 <- drm(no / total ~ dose, weights = total, curveid = time,
                   data = daphnids, fct = LL.2(), type = "binomial")
ED(daphnids.m1, 50)
```

## Model selection and averaging

`drc` selects among candidates by information criterion rather than averaging
over them.

```{r}
#| label: mselect
mselect(model_LL4, list(LL.3(), W1.4(), W2.4()))
```

`drc` does provide `maED()` for model-averaged effect concentrations, using
Akaike weights and a Buckland unconditional interval. It is worth knowing two
things about it before relying on it. It sums without removing missing values,
so a single candidate whose root-finder fails renders the whole average missing.
And it forms its weights as `exp(-AIC/2)` without centring on the minimum AIC,
which underflows to zero for every candidate once AIC reaches a few thousand, as
it does on individual-level data.

This is a genuine difference in philosophy rather than only in implementation.
`bayesnec` averages by default over a candidate set; `drc` selects by default,
and the selection is itself a source of uncertainty that the reported interval
does not include.

## Comparing groups

`EDcomp()` compares effect concentrations between curves fitted with `curveid`:

```{r}
#| label: edcomp
EDcomp(daphnids.m1, c(50, 50))
```

This is the frequentist counterpart to comparing posteriors, which module 7
covers. It returns a ratio with a confidence interval and a p-value, rather than
a probability that one estimate is lower than the other.

## Where the engines agree and where they differ

| | `drc` | `bayesnec` |
|---|---|---|
| Estimation | maximum likelihood | Bayesian, HMC via Stan |
| Interval | delta-method confidence interval | credible interval from the posterior |
| Threshold estimate | none | *NEC* from threshold equations |
| No-effect estimate | *NSEC* via `toxval` | *NEC* or *NSEC* |
| ECx reference | `f(0)`, the fitted response at dose zero | predicted mean at the lowest observed concentration |
| ECx default | percentage of the fitted **range** | percentage decline from control to zero |
| Candidate set | select one by AIC | average over the set by weight |
| Group comparison | ratio, confidence interval, p-value | probability of difference from the posteriors |

Four differences are worth stating in full.

**A confidence interval and a credible interval are different quantities.** They
are often of similar width and they do not mean the same thing. Label which one
a table reports.

**The control is a fitted value in `drc`.** `ED(reference = "control")` uses
`f(0)`, the value the fitted curve takes at dose zero, whether or not dose zero
was tested. Where the lowest tested concentration is well above zero, or the
curve is still declining there, that is an extrapolation, and the ECx inherits
it. `bayesnec` reads the control from the predicted mean at the lowest observed
concentration instead. On a well-behaved curve with a flat lower end the two
agree closely; on the shipped example above they agree to four decimal places.

**The distributional assumptions differ, and so will the estimates.** For a
continuous positive response `drc` will commonly be fitted Gaussian where
`bayesnec` chooses a gamma. The Gaussian fit is influenced more by the
high-response observations. Neither is an error, but the two engines will not
return the same point estimate, and a difference between them is not by itself
evidence that one is wrong.

**`drc`'s intervals can be too narrow, or absent.** With `type = "Poisson"` the
variance is fixed equal to the mean, so an over-dispersed count response gets
intervals narrower than the data support, by roughly the square root of the
dispersion. The remedy is to report the dispersion alongside the interval rather
than to rescale it, since changing the variance assumption is a change to the
model. Separately, the delta-method standard error is `NaN` wherever the variance
calculation goes negative, which is not rare and not confined to poor fits; each
one arrives as its own `"NaNs produced"` warning. Where standard errors feed a
model average, dropping candidates as they become unavailable silently changes
what the average is an average of, so fix the candidate set for the whole curve
rather than point by point.

## When to use which

For a well-determined monotonic curve with adequate replication and a response
that suits the same distribution in both, the two engines give much the same
answer, and `drc` gets there in seconds rather than minutes.

The Bayesian route earns its overhead where the model is harder: where priors are
needed to make a non-linear fit stable, where the candidate set genuinely does
not identify one curve, where a threshold estimate is wanted, or where the
posterior itself is the thing needed — to propagate uncertainty into a species
sensitivity distribution, or to state the probability that one substance is more
toxic than another.

## Provenance

```{r}
#| label: versions
#| comment: ""
for (p in c("drc", "toxval", "bayesnec")) {
  cat(sprintf("%-10s %s\n", p, as.character(packageVersion(p))))
}
cat(sprintf("%-10s %s\n", "R", getRversion()))
```

Every figure and number on this page is produced when it is built.

## References

::: {#refs}
:::
