---
title: "Fitting the same models with `drc`"
subtitle: "A frequentist engine asked the same questions, and where its answers differ"
bibliography: bayesnec.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(drc)
library(bayesnec)
library(ggplot2)
options(mc.cores = 4)

# Fitted curves are drawn from a prediction grid rather than by drc's own plot
# method, so that every figure on the page is built the same way and the code
# stays readable.
curve_data <- function(fit, from, to, n = 200) {
  grid <- data.frame(x = exp(seq(log(from), log(to), length.out = n)))
  grid$fit <- suppressWarnings(predict(fit, grid))
  grid
}
```

## Scope

This page is supplied for interest and is not taught.

`drc` [@Ritz2015] fits concentration-response models by maximum likelihood. It
is mature, widely used, and contains a great deal more than is shown here. What
follows asks it the questions the course asks of `bayesnec`, and sets out where
the two engines answer the same question and where they answer different ones.

Better sources exist for `drc` itself:

- Christian Ritz's [dose-response site](https://bioassay.dk/)
- the `drc` paper in *PLoS ONE* and its supplement of worked examples
  [@Ritz2015]
- community material at [rstats4ag.org](https://rstats4ag.org/dose-response-curves.html)

## Fitting a curve

`drm()` takes a formula, the data, and a mean function through `fct`. This is
the dataset from module 2, fitted with a four-parameter log-logistic curve.

```{r}
#| label: fit-ll4
data(nec_data)
model_LL4 <- drm(y ~ x, data = nec_data, fct = LL.4())
```

```{r}
#| label: fig-ll4
#| fig-height: 3.6
#| fig-cap: "A four-parameter log-logistic curve fitted to `nec_data`."
ggplot(nec_data, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_line(data = curve_data(model_LL4, min(nec_data$x), max(nec_data$x)),
            aes(x, fit), linewidth = 0.8, colour = "steelblue") +
  scale_x_continuous(trans = "log", breaks = c(0.1, 0.3, 1, 3)) +
  labs(x = "Concentration", y = "Response") +
  theme_classic()
```

The curve declines from about 0.9 at the lowest concentrations to near zero at
the highest, with the steep part between roughly 1 and 3.

## Reading the fit

```{r}
#| label: summary-ll4
summary(model_LL4)
```

A four-parameter `drc` model reports parameters named `b`, `c`, `d` and `e`:

| | meaning | `bayesnec` equivalent |
|---|---|---|
| `b` | slope, sometimes the hill slope | `beta` or `slope` |
| `c` | lower asymptote | `bot` |
| `d` | upper asymptote | `top` |
| `e` | inflection point | `ec50` |

The estimates above are a slope of 7.85, a lower asymptote indistinguishable
from zero, an upper asymptote of 0.897, and an inflection at 1.995. Where the
asymptotes span the full possible response range, `e` is the EC50, and that
holds here because the lower asymptote is zero.

Each parameter comes with a test of whether it differs from zero. That test is
seldom the question of interest. A lower asymptote indistinguishable from zero,
as here, is a reason to consider a three-parameter model with the lower limit
fixed, not evidence about toxicity.

Residuals can be examined for a continuous response, though not for binomial or
count data:

```{r}
#| label: fig-resid
#| fig-height: 3
ggplot(data.frame(r = residuals(model_LL4)), aes(r)) +
  geom_histogram(bins = 20, fill = "grey80", colour = "white") +
  labs(x = "Residual", y = "Count") +
  theme_classic()
shapiro.test(residuals(model_LL4))
```

The residuals are roughly symmetric about zero and the Shapiro-Wilk test returns
a p-value of 0.55, so there is no evidence against normality for this fit.

## Threshold models

`drc` provides threshold mean functions as well as smooth ones: `NEC.2`, `NEC.3`
and `NEC.4`. These estimate a breakpoint directly, in the same sense as the
`nec` equations of `bayesnec`, so a *NEC* is available from either engine.

```{r}
#| label: fit-nec3
model_NEC3 <- drm(y ~ x, data = nec_data, fct = NEC.3())
summary(model_NEC3)
```

The threshold is the `e` parameter, estimated at 1.523, with `d` the level of
the response below it.

```{r}
#| label: nec-ci
confint(model_NEC3)["e:(Intercept)", ]
```

Module 2 fitted `nec3param` to these same data with `bayesnec` and reported a
*NEC* of 1.54 with a 95% credible interval of 1.50 to 1.57. The `drc` threshold
above is 1.52 with a confidence interval of 1.49 to 1.55. The two engines are
estimating the same quantity here and they agree to within about one per cent.

```{r}
#| label: fig-nec3
#| fig-height: 3.6
#| fig-cap: "`NEC.3` fitted to the same data. The dashed line is the estimated threshold."
ggplot(nec_data, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_line(data = curve_data(model_NEC3, min(nec_data$x), max(nec_data$x)),
            aes(x, fit), linewidth = 0.8, colour = "firebrick") +
  geom_vline(xintercept = coef(model_NEC3)[["e:(Intercept)"]],
             linetype = 2, colour = "firebrick") +
  scale_x_continuous(trans = "log", breaks = c(0.1, 0.3, 1, 3)) +
  labs(x = "Concentration", y = "Response") +
  theme_classic()
```

The fitted curve holds flat at 0.890 until the threshold and declines
exponentially after it, which is the shape `nec3param` also fits.

## Effect concentrations

`ED()` returns effect concentrations, and two of its arguments determine what is
computed.

```{r}
#| label: ed
ED(model_LL4, c(10, 50))
```

### The basis of the percentage

For `type = "relative"`, which is the default, the percentage is a fraction of
the fitted response range, measured from the upper asymptote `d` down to the
lower asymptote `c`. This can be checked directly against the curve:

```{r}
#| label: range-check
cf <- coef(model_LL4)
c_lo <- cf[["c:(Intercept)"]]
d_hi <- cf[["d:(Intercept)"]]

ec10 <- ED(model_LL4, 10, display = FALSE)[1, 1]
ec50 <- ED(model_LL4, 50, display = FALSE)[1, 1]

c(response_at_EC10 = predict(model_LL4, data.frame(x = ec10)),
  d_minus_10pc_of_range = d_hi - 0.10 * (d_hi - c_lo),
  response_at_EC50 = predict(model_LL4, data.frame(x = ec50)),
  midpoint_of_range = c_lo + (d_hi - c_lo) / 2)
```

The response at `ED10` is 0.8076, which is exactly `d` minus a tenth of the
range, and the response at `ED50` is 0.4487, exactly the midpoint. The
percentage is therefore a percentage of `d − c`.

Both ends of that range are fitted parameters. `d` is the level the curve
approaches as concentration falls, whether or not anything was measured there,
and `c` is the level it approaches as concentration rises, which may lie beyond
the highest concentration tested. A design that does not reach either plateau
estimates that end by extrapolation, and every ECx inherits it.

`bayesnec` anchors differently. Its `ecx()` measures the decline from the
control, defined as the predicted mean at the lowest observed concentration,
which is a point inside the data rather than a fitted limit.

### The `reference` argument

The name suggests `reference` chooses the anchor. It does not. Its only effect
is a direction convention for increasing curves:

```{r}
#| label: reference-test
increasing <- transform(nec_data, y = 1 - y)
model_up <- drm(y ~ x, data = increasing, fct = LL.4())

rbind(
  decreasing = c(control = ED(model_LL4, 10, reference = "control", display = FALSE)[1, 1],
                 upper   = ED(model_LL4, 10, reference = "upper",   display = FALSE)[1, 1]),
  increasing = c(control = ED(model_up,  10, reference = "control", display = FALSE)[1, 1],
                 upper   = ED(model_up,  10, reference = "upper",   display = FALSE)[1, 1]))
```

On the decreasing curve both settings return 1.508. On the increasing curve they
return 1.511 and 2.494. For an ordinary concentration-response curve, which
declines, changing `reference` changes nothing.

### Values of `type` in each package

`bayesnec::ecx()` also takes a `type` argument and its values do not line up
with `drc`'s.

| `drc` `type` | what it means | nearest `bayesnec` `type` |
|---|---|---|
| `"relative"` (default) | percentage of the fitted range `d − c` | `"range"` |
| `"absolute"` | an absolute response value | `"direct"` |
| — | percentage decline from the control to zero | `"absolute"` (default) |

`drc`'s `"absolute"` asks for the concentration at which the response equals the
number supplied, so `ED(model_LL4, 10, type = "absolute")` asks where the
response equals 10. This response is bounded near 1, so no such concentration
exists and the call fails:

```{r}
#| label: ed-absolute
#| error: true
ED(model_LL4, 10, type = "absolute", display = FALSE)
```

The error is the correct behaviour. It is shown because the same word is the
default in `bayesnec` and means a percentage there, so translating
`ecx(fit, type = "absolute")` into `ED(m, 10, type = "absolute")` changes the
question being asked.

### Agreement between the engines

Where the fitted lower asymptote is zero and the curve is flat across the lowest
concentrations tested, the fitted range and the distance from the control to
zero are the same interval, and the two engines estimate the same quantity.
`nec_data` satisfies both conditions.

```{r}
#| label: fit-bnec-ecx4
#| message: false
#| warning: false
set.seed(333)
bnec_ecx4 <- bnec(y ~ crf(x, model = "ecx4param"), data = nec_data)
```

```{r}
#| label: compare-ecx
data.frame(
  quantity = c("EC10", "EC50"),
  drc_relative = c(ED(model_LL4, 10, display = FALSE)[1, 1],
                   ED(model_LL4, 50, display = FALSE)[1, 1]),
  bayesnec_range = c(ecx(bnec_ecx4, ecx_val = 10, type = "range")[1],
                     ecx(bnec_ecx4, ecx_val = 50, type = "range")[1]))
```

The EC50 estimates agree to three decimal places. The EC10 estimates differ by
about half a per cent, and the `drc` estimate falls inside the `bayesnec`
credible interval. Two different engines, fitting two differently parameterised
four-parameter curves by two different methods, return the same numbers to the
precision anyone would report.

### Divergence between the engines

Give the response a floor that is not zero and the fitted range is no longer the
distance from the control to zero. The two anchors then measure different
things.

```{r}
#| label: fit-floored
floored <- transform(nec_data, y = 0.3 + 0.6 * y)
model_floor <- drm(y ~ x, data = floored, fct = LL.4())

cf_f <- coef(model_floor)
c_f <- cf_f[["c:(Intercept)"]]
d_f <- cf_f[["d:(Intercept)"]]

# A 10% decline from the control, expressed as the percentage of the fitted
# range that drc needs, so that both quantities come from the same fit.
level_from_control <- 100 * (0.10 * d_f) / (d_f - c_f)

c(fitted_lower = c_f,
  fitted_upper = d_f,
  EC10_of_the_range = ED(model_floor, 10, display = FALSE)[1, 1],
  EC10_from_control = ED(model_floor, level_from_control, display = FALSE)[1, 1])
```

The fitted range runs from 0.344 to 0.838. A decline of ten per cent of that
range reaches a response of 0.789 and is met at a concentration of 1.511. A
decline of ten per cent from the control reaches 0.754 and is met at 1.620, some
seven per cent higher.

```{r}
#| label: fig-anchors
#| fig-height: 3.8
#| fig-cap: "The same fitted curve, with the two targets a ten per cent effect can mean. The upper dashed line is ten per cent of the fitted range below the control; the lower is ten per cent of the control."
gr <- curve_data(model_floor, min(floored$x), max(floored$x))
ec_range <- ED(model_floor, 10, display = FALSE)[1, 1]
ec_ctrl  <- ED(model_floor, level_from_control, display = FALSE)[1, 1]

ggplot(floored, aes(x, y)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_line(data = gr, aes(x, fit), linewidth = 0.8, colour = "steelblue") +
  geom_hline(yintercept = d_f - 0.10 * (d_f - c_f), linetype = 2, colour = "grey40") +
  geom_hline(yintercept = d_f - 0.10 * d_f, linetype = 2, colour = "firebrick") +
  geom_vline(xintercept = ec_range, linetype = 3, colour = "grey40") +
  geom_vline(xintercept = ec_ctrl,  linetype = 3, colour = "firebrick") +
  annotate("text", x = min(floored$x), y = c_f + 0.01,
           label = "fitted lower asymptote", hjust = 0, size = 3, colour = "grey40") +
  scale_x_continuous(trans = "log", breaks = c(0.1, 0.3, 1, 3)) +
  labs(x = "Concentration", y = "Response") +
  theme_classic()
```

Both estimates are correct, and they answer different questions. The
range-referenced EC10 is the concentration at which the response has fallen by a
tenth of the range the fitted curve spans, from its upper asymptote to its
lower. The control-referenced EC10 is the concentration at which the response has
fallen to nine tenths of its control level. Where a guideline value is being derived the second is normally what is
meant, and it is what `bayesnec` returns by default.

The practical rule when comparing an `ED()` result against an `ecx()` result is
to establish that both are anchored the same way, and to record in the write-up
which convention was used.

## A no-significant-effect concentration

The *NSEC* is defined from the fitted curve and the variability of the control
rather than from a model parameter [@fisherfox2023], so it can be derived from a
frequentist fit. Take the control level and a lower bound on it, then find the
concentration at which the curve crosses that bound.

For the threshold model the control level is the `d` parameter, and its standard
error is reported in the summary.

```{r}
#| label: nsec-derive
sig_val <- 0.01
pars <- summary(model_NEC3)$coefficients
d_est <- pars["d:(Intercept)", "Estimate"]
d_se  <- pars["d:(Intercept)", "Std. Error"]

# One-sided lower bound on the control at the 1 - sig_val level. This is the
# frequentist counterpart of the control posterior quantile bayesnec uses.
lower_bound <- d_est - qnorm(1 - sig_val) * d_se

crossing <- function(z) predict(model_NEC3, data.frame(x = z)) - lower_bound
nsec <- uniroot(crossing, interval = range(nec_data$x))$root

c(control = d_est, control_se = d_se, lower_bound = lower_bound, NSEC = nsec)
```

The control is estimated at 0.8902 with a standard error of 0.0049, giving a
lower 99 per cent bound of 0.8789. The curve crosses that bound at 1.531, which
is the *NSEC*. The threshold estimated by the same model is 1.523, slightly
lower, because the curve has to decline a little past the threshold before it
falls below the control's lower bound.

```{r}
#| label: fig-nsec
#| fig-height: 3.8
#| fig-cap: "The derived NSEC. The horizontal line is the lower bound on the control and the vertical line marks where the curve crosses it."
ggplot(nec_data, aes(x, y)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_line(data = curve_data(model_NEC3, min(nec_data$x), max(nec_data$x)),
            aes(x, fit), linewidth = 0.8, colour = "firebrick") +
  geom_hline(yintercept = lower_bound, linetype = 2, colour = "grey30") +
  geom_vline(xintercept = nsec, linetype = 3, colour = "grey30") +
  scale_x_continuous(trans = "log", breaks = c(0.1, 0.3, 1, 3)) +
  labs(x = "Concentration", y = "Response") +
  theme_classic()
```

The derivation above reads the control's uncertainty off the `d` parameter
because `predict(model_NEC3, se.fit = TRUE)` returns no standard error for this
fit. That is one instance of a wider problem with `drc` standard errors,
described at the end of this page.

Two things separate this from the `bayesnec` *NSEC*. The bound here is a normal
approximation from one standard error, where `bayesnec` takes a quantile of the
control posterior. And there is no interval on the result, because propagating
the uncertainty in both the control and the curve through the crossing would
require the delta method or a bootstrap.

## Model averaging

`drc` selects among candidates by information criterion:

```{r}
#| label: mselect
mselect(model_LL4, list(LL.3(), W1.4(), W2.4()))
```

`W2.4` has the lowest information criterion at −338.2, followed by `LL.3` at
−320.6.

It also averages, through `maED()`, using Akaike weights and a Buckland
unconditional interval:

```{r}
#| label: maed
maED(model_LL4, list(LL.3(), LL.4(), W1.4(), W2.4()), 10, display = FALSE)
```

The averaged EC10 is 1.565. The four candidates individually give 1.508, 1.508,
1.462 and 1.565, so the average sits at the value from `W2.4`, which took
almost all of the Akaike weight because its information criterion is nearly
twenty units below the others.

Two properties of `maED()` are visible in `drc::maED` itself and matter before
relying on it. It takes `na.rm = FALSE` and threads that through every `sum()`,
so one candidate whose root-finder fails renders every averaged value missing
rather than being dropped. And it forms weights as `exp(-AIC/2)` without
centring on the minimum AIC, so once AIC reaches a few thousand every term
underflows to zero in double precision and the weights become `0/0`. Centring on
the minimum resolves this and changes nothing where the uncentred form works.

The difference from `bayesnec` is one of default rather than of capability.
`bayesnec` averages unless asked to fit a single equation; `drc` fits one
equation unless asked to average, and the uncertainty in which equation was
selected does not appear in the interval of a single selected fit.

## Comparing groups

`EDcomp()` compares effect concentrations between curves fitted with `curveid`:

```{r}
#| label: binom
data(daphnids)
daphnids.m1 <- drm(no / total ~ dose, weights = total, curveid = time,
                   data = daphnids, fct = LL.2(), type = "binomial")
EDcomp(daphnids.m1, c(50, 50))
```

The ratio of the two EC50 values is 3.40 with a p-value of 0.003, so the 24-hour
and 48-hour curves differ. `bayesnec` answers the same question by comparing
posteriors and returns a probability that one estimate is lower than the other,
which module 7 covers.

## Agreement and divergence

| | `drc` | `bayesnec` |
|---|---|---|
| Estimation | maximum likelihood | Bayesian, HMC via Stan |
| Interval | confidence interval, delta method | credible interval from the posterior |
| Threshold equations | `NEC.2`, `NEC.3`, `NEC.4` | the ten `nec` equations |
| *NSEC* | derived by hand, as above | `nsec()`, with an interval |
| ECx anchor | the fitted range, `d` to `c` | the control, at the lowest observed concentration |
| Candidate set | one selected by AIC, or averaged with `maED()` | averaged by default |
| Group comparison | ratio, confidence interval, p-value | probability of difference from the posteriors |

A confidence interval and a credible interval are not the same quantity. A
credible interval states where the parameter lies with a given probability,
given the data, the model and the priors. A confidence interval states that the
procedure producing it would contain the true value in that proportion of
repeated experiments. The sentence a risk assessment usually wants — that there
is a 95 per cent probability the *NEC* lies between two values — follows from
the first and not from the second. Both are abbreviated CI, and `brms` prints
`l-95% CI`, so the interval type is worth naming in a results table.

The distributional assumptions differ and the point estimates will differ with
them. A continuous positive response will commonly be fitted Gaussian in `drc`
where `bayesnec` selects a gamma, and the Gaussian fit is influenced more by the
high-response observations. A difference between the engines is not by itself
evidence that either is wrong.

`drc` intervals can be too narrow, or missing. Under `type = "Poisson"` the
variance is fixed equal to the mean, so an over-dispersed count response gets
intervals narrower than the data support by roughly the square root of the
dispersion; report the dispersion alongside the interval rather than rescaling
it, since changing the variance assumption changes the model. Separately, the
delta-method standard error is `NaN` wherever the variance calculation goes
negative, which is neither rare nor confined to poor fits, and each occurrence
arrives as its own `"NaNs produced"` warning. The NSEC derivation above met this
in a mild form. Where standard errors feed a model average, dropping candidates
as their errors become unavailable changes what the average is an average of, so
fix the candidate set for the whole curve rather than point by point.

## Choosing between them

For a well-determined monotonic curve with adequate replication, fitted with the
same distribution in both, the two engines return much the same answer, as the
EC50 comparison above shows, and `drc` returns it in seconds rather than
minutes.

The Bayesian route repays its overhead where the problem is harder: where priors
are needed to stabilise a non-linear fit, where the candidate set does not
identify one curve, where an interval is wanted on a derived quantity such as
the *NSEC*, or where the posterior itself is needed to propagate uncertainty
into a species sensitivity distribution or to state the probability that one
substance is more toxic than another.

## Provenance

```{r}
#| label: versions
#| comment: ""
for (p in c("drc", "bayesnec", "brms")) {
  cat(sprintf("%-10s %s\n", p, as.character(packageVersion(p))))
}
cat(sprintf("%-10s %s\n", "R", getRversion()))
```

Every number and figure on this page is produced when it is built.

## References

::: {#refs}
:::
